---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
options(width = 76)
knitr::opts_chunk$set(
  fig.path = "man/figures/README-",
  digits = 3,
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  dev.args = list(bg = "transparent"),
  dpi = 300,
  cache.lazy = FALSE,
  out.width = "90%",
  fig.align = "center",
  fig.width = 9,
  fig.height = 6
)
ggplot2::theme_set(ggplot2::theme_bw())
options(
  dplyr.print_min = 6,
  dplyr.print_max = 6,
  pillar.max_footer_lines = 2,
  pillar.min_chars = 15,
  stringr.view_n = 6,
  pillar.bold = TRUE,
  width = 77
)
```
```{r pkgs, include=FALSE, echo=FALSE}
library(epipredict)
library(epidatr)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(magrittr)
library(purrr)
library(scales)
```

```{r coloration, include=FALSE, echo=FALSE}
base <- "#002676"
primary <- "#941120"
secondary <- "#f9c80e"
tertiary <- "#177245"
fourth_colour <- "#A393BF"
fifth_colour <- "#2e8edd"
colvec <- c(
  base = base, primary = primary, secondary = secondary,
  tertiary = tertiary, fourth_colour = fourth_colour,
  fifth_colour = fifth_colour
)
library(epiprocess)
suppressMessages(library(tidyverse))
theme_update(legend.position = "bottom", legend.title = element_blank())
delphi_pal <- function(n) {
  if (n > 6L) warning("Not enough colors in this palette!")
  unname(colvec)[1:n]
}
scale_fill_delphi <- function(..., aesthetics = "fill") {
  discrete_scale(aesthetics = aesthetics, palette = delphi_pal, ...)
}
scale_color_delphi <- function(..., aesthetics = "color") {
  discrete_scale(aesthetics = aesthetics, palette = delphi_pal, ...)
}
scale_colour_delphi <- scale_color_delphi
```

# Epipredict

<!-- badges: start -->
[![R-CMD-check](https://github.com/cmu-delphi/epipredict/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/cmu-delphi/epipredict/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

[`{epipredict}`](https://cmu-delphi.github.io/epipredict/) is a framework for building transformation and forecasting pipelines for epidemiological and other panel time-series datasets.
In addition to tools for building forecasting pipelines, it contains a number of “canned” forecasters meant to run with little modification as an easy way to get started forecasting.

It is designed to work well with
[`{epiprocess}`](https://cmu-delphi.github.io/epiprocess/), a utility for time series handling and geographic processing in an epidemiological context.
Both of the packages are meant to work well with the panel data provided by
[`{epidatr}`](https://cmu-delphi.github.io/epidatr/).
Pre-compiled example datasets are also available in
[`{epidatasets}`](https://cmu-delphi.github.io/epidatasets/).

If you are looking for detail beyond the package documentation, see our
[forecasting book](https://cmu-delphi.github.io/delphi-tooling-book/).


## Installation

Unless you’re planning on contributing to package development, we suggest using the stable version.
To install, run:

```r
# Stable version
pak::pkg_install("cmu-delphi/epipredict@main")

# Dev version
pak::pkg_install("cmu-delphi/epipredict@dev")
```

The documentation for the stable version is at
<https://cmu-delphi.github.io/epipredict>, while the development version is at
<https://cmu-delphi.github.io/epipredict/dev>.


## Motivating example


<details>
<summary> Required packages </summary>

```{r install, run = FALSE}
library(epipredict)
library(epidatr)
library(epiprocess)
library(dplyr)
library(ggplot2)
```
</details>

To demonstrate using [`{epipredict}`](https://cmu-delphi.github.io/epipredict/) for forecasting, say we want to
predict COVID-19 deaths per 100k people for each of a subset of states

```{r subset_geos}
used_locations <- c("ca", "ma", "ny", "tx")
```

on

```{r fc_date}
forecast_date <- as.Date("2021-08-01")
```

We will be using a subset of
[Johns Hopkins Center for Systems Science and Engineering deaths data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html).
Below the fold, we pull the dataset from the epidata API and clean it.

<details>
<summary> Creating the dataset using `{epidatr}` and `{epiprocess}` </summary>

This section is intended to demonstrate some of the ubiquitous cleaning operations needed to be able to forecast.
A subset of the dataset prepared here is also included ready-to-go in [`{epipredict}`](https://cmu-delphi.github.io/epipredict/) as `covid_case_death_rates`.

First we pull both `jhu-csse` cases and deaths data from the
[Delphi API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html) using the
[`{epidatr}`](https://cmu-delphi.github.io/epidatr/) package:

```{r case_death, warning = FALSE, eval = TRUE}
cases <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_7dav_incidence_prop",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200601, 20211231),
  geo_values = "*"
) |>
  select(geo_value, time_value, case_rate = value)

deaths <- pub_covidcast(
  source = "jhu-csse",
  signals = "deaths_7dav_incidence_prop",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200601, 20211231),
  geo_values = "*"
) |>
  select(geo_value, time_value, death_rate = value)
cases_deaths <-
  full_join(cases, deaths, by = c("time_value", "geo_value")) |>
  filter(geo_value %in% used_locations) |>
  as_epi_df(as_of = as.Date("2022-01-01"))
```

Since visualizing the results on every geography is somewhat overwhelming,
we’ll only train on a subset of locations.

```{r date, warning = FALSE}
# plotting the data as it was downloaded
cases_deaths |>
  autoplot(
    case_rate,
    death_rate,
    .color_by = "none"
  ) +
  facet_grid(
    rows = vars(.response_name),
    cols = vars(geo_value),
    scale = "free"
  ) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y %b") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

As with the typical dataset, we will need to do some cleaning to
make it actually usable; we’ll use some utilities from
[`{epiprocess}`](https://cmu-delphi.github.io/epiprocess/) for this.
Specifically we'll trim outliers, especially negative values:

```{r outlier}
cases_deaths <-
  cases_deaths |>
  group_by(geo_value) |>
  mutate(
    outlr_death_rate = detect_outlr_rm(
      time_value, death_rate,
      detect_negatives = TRUE
    ),
    outlr_case_rate = detect_outlr_rm(
      time_value, case_rate,
      detect_negatives = TRUE
    )
  ) |>
  unnest(cols = starts_with("outlr"), names_sep = "_") |>
  ungroup() |>
  mutate(
    death_rate = outlr_death_rate_replacement,
    case_rate = outlr_case_rate_replacement
  ) |>
  select(geo_value, time_value, case_rate, death_rate)
```
</details>

After downloading and cleaning the cases and deaths data, we can plot
a subset of the states, marking the desired forecast date:

<details>
<summary> Plot </summary>

```{r plot_locs}
forecast_date_label <-
  tibble(
    geo_value = rep(used_locations, 2),
    .response_name = c(rep("case_rate", 4), rep("death_rate", 4)),
    dates = rep(forecast_date - 7 * 2, 2 * length(used_locations)),
    heights = c(rep(150, 4), rep(0.75, 4))
  )
processed_data_plot <-
  cases_deaths |>
  filter(geo_value %in% used_locations) |>
  autoplot(
    case_rate,
    death_rate,
    .color_by = "none"
  ) +
  facet_grid(
    rows = vars(.response_name),
    cols = vars(geo_value),
    scale = "free"
  ) +
  geom_vline(aes(xintercept = forecast_date)) +
  geom_text(
    data = forecast_date_label,
    aes(x = dates, label = "forecast\ndate", y = heights),
    size = 3, hjust = "right"
  ) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y %b") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
</details>
```{r show-processed-data, warning=FALSE, echo=FALSE}
processed_data_plot
```

To make a forecast, we will use a simple “canned” auto-regressive forecaster to
predict the death rate four weeks into the future using lagged[^3] deaths and
cases.

[^3]: lagged by 3 in this context meaning using the value from 3 days ago.

```{r make-forecasts, warning=FALSE}
four_week_ahead <- arx_forecaster(
  cases_deaths |> filter(time_value <= forecast_date),
  outcome = "death_rate",
  predictors = c("case_rate", "death_rate"),
  args_list = arx_args_list(
    lags = list(c(0, 1, 2, 3, 7, 14), c(0, 7, 14)),
    ahead = 4 * 7,
    quantile_levels = c(0.1, 0.25, 0.5, 0.75, 0.9)
  )
)
four_week_ahead
```

In our model setup, we are defining as predictors case rate lagged 0-3
days, one week, and two weeks, and death rate lagged 0-2 weeks.
The result `four_week_ahead` is both a fitted model object which could be used
any time in the future to create different forecasts, and a set of predicted
values (and prediction intervals) for each location 28 days after the forecast
date.

Plotting the prediction intervals on the true values for our location subset[^2]:

[^2]: Alternatively, you could call `autoplot(four_week_ahead, plot_data =
    cases_deaths)` to get the full collection of forecasts. This is too busy for
    the space we have for plotting here.

<details>
<summary> Plot </summary>
```{r plotting_forecast, warning=FALSE}
epiworkflow <- four_week_ahead$epi_workflow
restricted_predictions <-
  four_week_ahead$predictions |>
  rename(time_value = target_date, value = .pred) |>
  mutate(.response_name = "death_rate")
forecast_plot <-
  four_week_ahead |>
  autoplot(plot_data = cases_deaths) +
  geom_vline(aes(xintercept = forecast_date)) +
  geom_text(
    data = forecast_date_label %>% filter(.response_name == "death_rate"),
    aes(x = dates, label = "forecast\ndate", y = heights),
    size = 3, hjust = "right"
  ) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y %b") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
</details>

```{r show-single-forecast, warning=FALSE, echo=FALSE}
forecast_plot
```

And as a tibble of quantile level-value pairs:
```{r pivot_wider}
four_week_ahead$predictions |>
  select(-.pred) |>
  pivot_quantiles_longer(.pred_distn) |>
  select(geo_value, forecast_date, target_date, quantile = .pred_distn_quantile_level, value = .pred_distn_value)
```

The yellow dot gives the median prediction, while the blue intervals give the
25-75%, the 10-90%, and 2.5-97.5%  inter-quantile ranges[^4].
For this particular day and these locations, the forecasts are relatively
accurate, with the true data being at least within the 10-90% interval.
A couple of things to note:

1. `epipredict` methods are primarily direct forecasters; this means we don't need to
   predict 1, 2,..., 27 days ahead to then predict 28 days ahead.
2. All of our existing engines are geo-pooled, meaning the training data is
   shared across geographies. This has the advantage of increasing the amount of
   available training data, with the restriction that the data needs to be on
   comparable scales, such as rates.

## Getting Help
If you encounter a bug or have a feature request, feel free to file an [issue on
our GitHub page](https://github.com/cmu-delphi/epipredict/issues).
For other questions, feel free to reach out to the authors, either via this
[contact form](https://docs.google.com/forms/d/e/1FAIpQLScqgT1fKZr5VWBfsaSp-DNaN03aV6EoZU4YljIzHJ1Wl_zmtg/viewform),
email, or the InsightNet Slack.

[^4]: Note that these are not the same quantiles that we fit when creating
    `four_week_ahead`. They are extrapolated from those quantiles using `extrapolate_quantiles()` (which assumes an exponential decay in the tails).
