---
title: "Accurately backtesting forecasters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accurately backtesting forecasters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source(here::here("vignettes/_common.R"))
```

Backtesting is a crucial step in the development of forecasting models. It
involves testing the model on historical time periods to see how well it generalizes to new
data.

In the context of
epidemiological forecasting, to do backtesting accurately, we need to account
for the fact that the data available at _the time of the forecast_ would have been
different from the data available at the time of the _backtest_.
This is because
new data is constantly being collected and added to the dataset, and old data potentially revised.
Training and making
predictions only on finalized data can lead to overly optimistic estimates of accuracy
(see, for example, [McDonald et al.
2021](https://www.pnas.org/content/118/51/e2111453118/) and the references
therein).

In the `{epiprocess}` package, we provide the function `epix_slide()` to help conviently perform version-faithful forecasting by only using the data as
it would have been available at forecast reference time.
In this vignette, we will demonstrate how to use `epix_slide()` to backtest an
auto-regressive forecaster constructed using `arx_forecaster()` on historical
COVID-19 case data from the US and Canada. 

# Getting case data from US states into an `epi_archive`

```{r pkgs, message=FALSE}
# Setup
library(epipredict)
library(epiprocess)
library(epidatr)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(magrittr)
library(purrr)
library(lubridate)
```

First, we create an `epi_archive()` to store the version history of the
percentage of doctor's visits with CLI (COVID-like illness) computed from
medical insurance claims and the number of new confirmed COVID-19 cases per
100,000 population (daily) for 4 states 

```{r grab-epi-data}
# Select the `percent_cli` column from the data archive
doctor_visits <- archive_cases_dv_subset$DT |>
  select(geo_value, time_value, version, percent_cli) |>
  tidyr::drop_na(percent_cli) |>
  as_epi_archive(compactify = TRUE)
```

The data can also be fetched from the Delphi Epidata API with the following
query:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(epidatr)
doctor_visits <- pub_covidcast(
  source = "doctor-visits",
  signals = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
) |>
  # The version date column is called `issue` in the Epidata API. Rename it.
  select(version = issue, geo_value, time_value, percent_cli = value) |>
  as_epi_archive(compactify = TRUE)
```

In the interest of computational speed, we limit the dataset to 4 states and 
2020–2021, but the full archive can be used in the same way and has performed
well in the past.

We choose this dataset in particular partly because it is revision heavy; for
example, here is a plot that compares monthly snapshots of the data.

<details>
<summary>Code for plotting</summary>
```{r plot_revision_example, warn = FALSE, message = FALSE}
geo_choose <- "ca"
forecast_dates <- seq(from = as.Date("2020-08-01"), to = as.Date("2021-11-01"), by = "1 month")
percent_cli_data <- bind_rows(
  # Snapshotted data for the version-faithful forecasts
  map(
    forecast_dates,
    ~ doctor_visits |>
      epix_as_of(.x) |>
      mutate(version = .x)
  ) |>
    bind_rows() |>
    mutate(version_faithful = "Version faithful"),
  # Latest data for the version-un-faithful forecasts
  doctor_visits |>
    epix_as_of(doctor_visits$versions_end) |>
    mutate(version_faithful = "Version un-faithful")
) |> as_tibble()
p0 <- autoplot(
  archive_cases_dv_subset, percent_cli, 
  .versions = seq(ymd("2020-07-01"), ymd("2021-11-30"), by = "1 month"), 
  .mark_versions = TRUE,
  .facet_filter = (geo_value == "ca")
) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %Y") +
  labs(x = "", y = "% of doctor's visits with\n Covid-like illness") + 
  scale_color_viridis_c(option = "turbo", guide = guide_legend(reverse=TRUE), direction = -1) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(c(0, 0.05))) +
  theme(legend.position = "none")
```
</details>

```{r plot_just_revisioning, echo = FALSE, warning = FALSE, warn = FALSE, message = FALSE}
p0
```

The snapshots are taken on the first of each month, with the vertical dashed
line representing the issue date for the time series of the corresponding
color.
For example, the snapshot on March 1st, 2021 is aquamarine, and increases to
slightly over 10.
Every series is necessarily to the left of the snapshot date (since all known
values must happen before the snapshot is taken[^4]).
The grey line overlaying the various snapshots represents the "final
value", which is just the snapshot at the last version in the archive (the
`versions_end`).

Comparing with the grey line tells us how much the value at the time of the
snapshot differs with what was eventually reported.
The drop in January 2021 in the snapshot on `2021-02-01` was initially reported
as much steeper than it eventually turned out to be, while in the period after
that the values were initially reported as higher than they actually were.

Handling data latency is important in both real-time forecasting and retrospective
forecasting.
Looking at the very first snapshot, `2020-08-01` (the red dotted
vertical line), there is a noticeable gap between the forecast date and the end
of the red time-series to its left.
In fact, if we take a snapshot and get the last `time_value`,

```{r}
doctor_visits |>
  epix_as_of(as.Date("2020-08-01")) |>
  pull(time_value) |>
  max()
```

the last day of data is the 25th, a entire week before `2020-08-01`.
This can require some effort to work around, especially if the latency is
variable; see `step_adjust_latency()` for some methods included in this package.
Much of that functionality is built into `arx_forecaster()` using the parameter
`adjust_ahead`, which we will use below.


# Backtesting a simple autoregressive forecaster

In addition to outlier detection and nowcasting, a common use case of `epiprocess::epi_archive()` object is for accurate model
back-testing.

To start, let's use a simple autoregressive forecaster to predict `percent_cli`, the percentage
of doctor's hospital visits associated with COVID-like illness, 14
days in the future. 
For increased accuracy we will use quantile regression.

## Comparing a single day and ahead

As a sanity check before we backtest the _entire_ dataset, let's 
forecast a single day in the middle of the dataset. 
We can do this by setting the `.versions` argument in `epix_slide()`:

```{r single_version, warn = FALSE}
forecast_date <- as.Date("2021-04-06")
forecasts <- doctor_visits |>
  epix_slide(
    ~ arx_forecaster(
      .x,
      outcome = "percent_cli",
      predictors = "percent_cli",
      args_list = arx_args_list()
    )$predictions |>
      pivot_quantiles_wider(.pred_distn),
    .versions = forecast_date
  )
```

We need truth data to compare our forecast against. We can construct it by using `epix_as_of()` to snapshot
the archive at the last available date[^1].

_Note:_ We always want to compare our forecasts to actual (most recently reported) values because that is the outcome we care about.
`as_of` data is useful for understanding why we're getting the forecasts we're getting, but `as_of` values are only preliminary outcomes.
Therefore, it's not meaningful to use them for evaluating the performance of a forecast.
Unfortunately, it's not uncommon for revisions to cause poor (final) performance of a forecaster that was decent at the time of the forecast.

```{r compare_single_with_result}
forecasts |>
  inner_join(
    doctor_visits |>
      epix_as_of(doctor_visits$versions_end),
    by = c("geo_value", "target_date" = "time_value")
  ) |>
  select(geo_value, forecast_date, .pred, `0.05`, `0.95`, percent_cli)
```

`.pred` corresponds to the point forecast (median), and `0.05` and `0.95`
correspond to the 5th and 95th quantiles.
The `percent_cli` truth data falls within the prediction intervals, so our
implementation passes a simple validation.

## Comparing version faithful and version un-faithful forecasts

Now let's compare the behavior of this forecaster, both properly considering data versioning 
("version faithful") and ignoring data versions ("version un-faithful").

For the version un-faithful approach, we need to do some setup if we want to use `epix_slide` for backtesting.
We want to simulate a data set that receives finalized updates every day, that is, a data set with no revisions.
To do this, we will snapshot the latest version of the data to create a synthetic data set, and convert it into an archive
where `version = time_value`[^2].

```{r}
archive_cases_dv_subset_faux <- doctor_visits |>
  epix_as_of(doctor_visits$versions_end) |>
  mutate(version = time_value) |>
  as_epi_archive()
```

For the version faithful approach, we will continue using the original `epi_archive` object containing all version updates.

We will also create the helper function `forecast_wrapper()` to let us easily map across aheads.

```{r arx-kweek-preliminaries, warning = FALSE}
forecast_wrapper <- function(
    epi_data, aheads, outcome, predictors, process_data = identity
    ) {
  map(
    aheads,
    \(ahead) {
      arx_forecaster(
        process_data(epi_data), outcome, predictors,
        args_list = arx_args_list(
          ahead = ahead,
          lags = c(0:7, 14, 21),
          adjust_latency = "extend_ahead"
        )
      )$predictions |> pivot_quantiles_wider(.pred_distn)
    }
  ) |> bind_rows()
}
```

_Note:_ In the helper function, we're using the parameter `adjust_latency`.
We need to use it because the most recently released data may still be several days old on any given forecast date (lag > 0); 
`adjust_latency` will modify the forecaster to compensate[^5].
See the function `step_adjust_latency()` for more details and examples.

Now that we're set up, we can generate forecasts for both the version faithful and un-faithful
archives, and bind the results together.

```{r generate_forecasts, warning = FALSE}
forecast_dates <- seq(
  from = as.Date("2020-09-01"),
  to = as.Date("2021-11-01"),
  by = "1 month"
)
aheads <- c(1, 7, 14, 21, 28)

version_unfaithful <- archive_cases_dv_subset_faux |>
  epix_slide(
    ~ forecast_wrapper(.x, aheads, "percent_cli", "percent_cli"),
    .before = 120,
    .versions = forecast_dates
  ) |>
  mutate(version_faithful = "Version un-faithful")

version_faithful <- doctor_visits |>
  epix_slide(
    ~ forecast_wrapper(.x, aheads, "percent_cli", "percent_cli"),
    .before = 120,
    .versions = forecast_dates
  ) |>
  mutate(version_faithful = "Version faithful")

forecasts <-
  bind_rows(
    version_unfaithful,
    version_faithful
  )
```

`arx_forecaster()` does all the heavy lifting.
It creates and lags copies of the features (here, the response and doctors visits), 
creates and leads copies of the target while respecting timestamps and locations, fits a
forecasting model using the specified engine, creates predictions, and
creates non-parametric confidence bands.

To see how the version faithful and un-faithful predictions compare, let's plot them on top of the latest case
rates, using the same versioned plotting method as above.
Note that even though we fit the model on four states (California, Texas, Florida, and
New York), we'll just display the results for two states, California (CA) and Florida
(FL), to get a sense of the model performance while keeping the graphic simple.

<details>
<summary>Code for plotting</summary>

```{r plot_ca_forecasts, warning = FALSE}
geo_choose <- "ca"
forecasts_filtered <- forecasts |>
  filter(geo_value == geo_choose) |>
  mutate(time_value = version)
# we need to add the ground truth data to the version faithful plot as well
plotting_data <- bind_rows(
  percent_cli_data,
  percent_cli_data |>
    filter(version_faithful == "Version un-faithful") |>
    mutate(version = max(percent_cli_data$version)) |>
    mutate(version_faithful = "Version faithful")
)
p1 <- ggplot(data = forecasts_filtered, aes(x = target_date, group = time_value)) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = (time_value)), alpha = 0.4) +
  geom_line(aes(y = .pred, color = (time_value)), linetype = 2L) +
  geom_point(aes(y = .pred, color = (time_value)), size = 0.75) +
  # the forecast date
  geom_vline(
    data = percent_cli_data |> filter(geo_value == geo_choose) |> select(-version_faithful),
    aes(color = version, xintercept = version, group = version),
    lty = 2
  ) +
  # the underlying data
  geom_line(
    data = plotting_data |> filter(geo_value == geo_choose),
    aes(x = time_value, y = percent_cli, color = (version), group = version),
    inherit.aes = FALSE, na.rm = TRUE
  ) +
  facet_grid(version_faithful ~ geo_value, scales = "free") +
  scale_x_date(breaks = "2 months", date_labels = "%b %Y") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  labs(x = "Date", y = "smoothed, day of week adjusted covid-like doctors visits") +
  scale_color_viridis_c(option = "turbo", direction = -1) +
  scale_fill_viridis_c(option = "turbo", direction = -1) +
  theme(legend.position = "none")
p1
```

```{r plot_fl_forecasts, warning = FALSE}
geo_choose <- "fl"
forecasts_filtered <- forecasts |>
  filter(geo_value == geo_choose) |>
  mutate(time_value = version)

forecasts_filtered %>% names
p2 <-
  ggplot(data = forecasts_filtered, aes(x = target_date, group = time_value)) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = (time_value)), alpha = 0.4) +
  geom_line(aes(y = .pred, color = (time_value)), linetype = 2L) +
  geom_point(aes(y = .pred, color = (time_value)), size = 0.75) +
  # the forecast date
  geom_vline(
    data = percent_cli_data |> filter(geo_value == geo_choose) |> select(-version_faithful),
    aes(color = version, xintercept = version, group = version),
    lty = 2
  ) +
  # the underlying data
  geom_line(
    data = plotting_data |> filter(geo_value == geo_choose),
    aes(x = time_value, y = percent_cli, color = (version), group = version),
    inherit.aes = FALSE, na.rm = TRUE
  ) +
  facet_grid(version_faithful ~ geo_value, scales = "free") +
  scale_x_date(breaks = "2 months", date_labels = "%b %Y") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  labs(x = "Date", y = "smoothed, day of week adjusted covid-like doctors visits") +
  scale_color_viridis_c(option = "turbo", direction = -1) +
  scale_fill_viridis_c(option = "turbo", direction = -1) +
  theme(legend.position = "none")
p2
```
</details>

```{r show-plot1, warning = FALSE, echo=FALSE}
p1
```

The version faithful and un-faithful forecasts look moderately similar except for the 1 day horizons
(although neither approach produces amazingly accurate forecasts).

In the version faithful case for California, the March 2021 forecast (turquoise)
starts at a value just above 10, which is very well lined up with reported values leading up to that forecast.
The measured and forecasted trends are also concordant (both increasingly moderately fast).

Because the data for this time period was later adjusted down with a decreasing trend, the March 2021 forecast looks quite bad compared to finalized data.

The equivalent version un-faithful forecast starts at a value of 5, which is in line with the finalized data but would have been out of place compared to the version data.

```{r show-plot2, warning = FALSE, echo=FALSE}
p2
```

Now let's look at Florida.
In the version faithful case, the three late-2021 forecasts (purples and pinks) starting in September predict very low values, near 0.
The trend leading up to each forecast shows a substantial decrease, so these forecasts seem appropriate and we would expect them to score fairly well on various performance metrics when compared to the versioned data.

However in hindsight, we know that early versions of the data systematically under-reported COVID-related doctor visits such that these forecasts don't actually perform well compared to _finalized_ data.
In this example, version faithful forecasts predicted values at or near 0 while finalized data shows values in the 5-10 range.
As a result, the version un-faithful forecasts for these same dates are quite a bit higher, and would perform well when scored using the finalized data and poorly with versioned data.

In general, the longer ago a forecast was made, the worse its performance is compared to finalized data. Finalized data accumulates revisions over time that make it deviate more and more from the non-finalized data a model was trained on.
Forecasts _trained_ on finalized data will of course appear to perform better when _scored_ on finalized data, but will have unknown performance on the non-finalized data we need to use if we want timely predictions.

Without using data that would have been available on the actual forecast date, 
you have little insight into what level of performance you
can expect in practice.

Good performance of a version un-faithful model is a mirage; it is only achievable if the training data has no revisions.
If a data source has any revisions, version un-faithful-level performance is unachievable when making forecasts in real time.


[^1]: For forecasting a single day like this, we could have actually just used
    `doctor_visits |> epix_as_of(forecast_date)` to get the relevant snapshot, and then fed that into `arx_forecaster()` as we did in the [landing
page](../index.html#motivating-example).


[^2]: Generally we advise against this; the only time to consider faking
 versioning like this are if you're back-testing data with no versions available
 at all, or if you're doing an explicit comparison like this. If you have no
 versions you should assume performance is worse than what the test would
 otherwise suggest.
    
[^4]: Until we have a time machine

[^5]: In this case by adjusting the length of the ahead so that it is actually
 forecasting from the last day of data (e.g. for 2 day latent data and a true
 ahead of 5, the `extended_ahead` would actually be 7)
