---
title: "Using epipredict on non-epidemic panel data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using epipredict on non-epidemic panel data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_common.R")
```

```{r libraries, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(parsnip)
library(recipes)
library(epiprocess)
library(epipredict)
library(epidatasets)
library(ggplot2)
theme_set(theme_bw())
```

[Panel data](https://en.wikipedia.org/wiki/Panel_data), or longitudinal data,
contain cross-sectional measurements of subjects over time. The `epipredict`
package is most suitable for running forecasters on epidemiological panel data.
An example of this is the [`covid_case_death_rates`](
  https://cmu-delphi.github.io/epidatasets/reference/covid_case_death_rates.html)
dataset, which contains daily state-wise measures of `case_rate` and 
`death_rate` for COVID-19 in 2021:

```{r epi-panel-ex, include=T}
head(covid_case_death_rates, 3)
```

`epipredict` functions work with data in
[`epi_df`](https://cmu-delphi.github.io/epiprocess/reference/epi_df.html)
format. Despite the stated goal and name of the package, other panel datasets
are also valid candidates for `epipredict` functionality, as long as they are
in `epi_df` format.

```{r employ-stats, include=F}
year_start <- min(grad_employ_subset$time_value)
year_end <- max(grad_employ_subset$time_value)
```

# Example panel data overview

In this vignette, we will demonstrate using `epipredict` with employment panel
data from Statistics Canada. We will be using
[
  Table 37-10-0115-01: Characteristics and median employment income of
  longitudinal cohorts of postsecondary graduates two and five years after
  graduation, by educational qualification and field of study (primary
  groupings)
](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3710011501).

The full dataset contains yearly median employment income two and five years
after graduation, and number of graduates. The data is stratified by
variables such as geographic region (Canadian province), education, and
age group. The year range of the dataset is `r year_start` to `r year_end`,
inclusive. The full dataset also contains metadata that describes the
quality of data collected. For demonstration purposes, we make the following
modifications to get a subset of the full dataset:

* Only keep provincial-level geographic region (the full data also has
"Canada" as a region)
* Only keep "good" or better quality data rows, as indicated by the [`STATUS`](
  https://www.statcan.gc.ca/en/concepts/definitions/guide-symbol) column
* Choose a subset of covariates and aggregate across the remaining ones. The
chosen covariates are age group, and educational qualification.

To use this data with `epipredict`, we need to convert it into `epi_df` format
using `epiprocess::as_epi_df()`
with additional keys. In our case, the additional keys are `age_group`,
and `edu_qual`. Note that in the above modifications, we encoded `time_value`
as type `integer`. This lets us set `time_type = "year"`, and ensures that
lag and ahead modifications later on are using the correct time units. See the
`epiprocess::epi_df` for
a list of all the `time_type`s available.

```{r data-dim, include=F}
employ_rowcount <- format(nrow(grad_employ_subset), big.mark = ",")
employ_colcount <- length(names(grad_employ_subset))
```

Now, we are ready to use `grad_employ_subset` with `epipredict`.
Our `epi_df` contains `r employ_rowcount` rows and `r employ_colcount` columns.
Here is a quick summary of the columns in our `epi_df`:

* `time_value` (time value): year in `date` format
* `geo_value` (geo value): province in Canada
* `num_graduates` (raw, time series value): number of graduates
* `med_income_2y` (raw, time series value): median employment income 2 years
after graduation
* `med_income_5y` (raw, time series value): median employment income 5 years
after graduation
* `age_group` (key): one of two age groups, either 15 to 34 years, or 35 to 64
years
* `edu_qual` (key): one of 32 unique educational qualifications, e.g.,
"Master's diploma"

```{r preview-data, include=T}
# Rename for simplicity
employ <- grad_employ_subset
sample_n(employ, 6)
```

In the following sections, we will go over pre-processing the data in the
`epi_recipe` framework, and fitting a model and making predictions within the
`epipredict` framework and using the package's canned forecasters.

# Autoregressive (AR) model to predict number of graduates in a year

## Pre-processing

As a simple example, let's work with the `num_graduates` column for now. We will
first pre-process by standardizing each numeric column by the total within
each group of keys. We do this since those raw numeric values will vary greatly
from province to province since there are large differences in population.

```{r employ-small, include=T}
employ_small <- employ %>%
  group_by(geo_value, age_group, edu_qual) %>%
  # Select groups where there are complete time series values
  filter(n() >= 6) %>%
  mutate(
    num_graduates_prop = num_graduates / sum(num_graduates),
    med_income_2y_prop = med_income_2y / sum(med_income_2y),
    med_income_5y_prop = med_income_5y / sum(med_income_5y)
  ) %>%
  ungroup()
head(employ_small)
```

Below is a visualization for a sample of the small data for British Columbia and Ontario.
Note that some groups
do not have any time series information since we filtered out all time series
with incomplete dates.

```{r employ-small-graph, include=T, eval=T, fig.width=9, fig.height=6}
employ_small %>%
  filter(geo_value %in% c("British Columbia", "Ontario")) %>%
  filter(grepl("degree", edu_qual, fixed = T)) %>%
  group_by(geo_value, time_value, edu_qual, age_group) %>%
  summarise(num_graduates_prop = sum(num_graduates_prop), .groups = "drop") %>%
  ggplot(aes(x = time_value, y = num_graduates_prop, color = geo_value)) +
  geom_line() +
  scale_colour_manual(values = c("Cornflowerblue", "Orange"), name = "") +
  facet_grid(rows = vars(edu_qual), cols = vars(age_group)) +
  xlab("Year") +
  ylab("Percentage of gratuates") +
  theme(legend.position = "bottom")
```

We will predict the standardized number of graduates (a proportion) in the
next year (time $t+1$) using an autoregressive model with three lags (i.e., an
AR(3) model). Such a model is represented algebraically like this:

\[
  y_{t+1,ijk} =
  \alpha_0 + \alpha_1 y_{tijk} + \alpha_2 y_{t-1,ijk} + \alpha_3 y_{t-2,ijk} + \epsilon_{tijk}
\]

where $y_{tij}$ is the proportion of graduates at time $t$ in location $i$ and
age group $j$ with education quality $k$.

In the pre-processing step, we need to create additional columns in `employ` for
each of $y_{t+1,ijk}$, $y_{tijk}$, $y_{t-1,ijk}$, and $y_{t-2,ijk}$.
We do this via an
`epi_recipe`. Note that creating an `epi_recipe` alone doesn't add these
outcome and predictor columns; the recipe just stores the instructions for
adding them.

Our `epi_recipe` should add one `ahead` column representing $y_{t+1,ijk}$ and
3 `lag` columns representing $y_{tijk}$, $y_{t-1,ijk}$, and $y_{t-2,ijk}$
(it's more accurate to think of the  0th "lag" as the "current" value with 2 lags,
but that's not quite how the processing works).
Also note that
since we specified our `time_type` to be `year`, our `lag` and `lead`
values are both in years.

```{r make-recipe, include=T, eval=T}
r <- epi_recipe(employ_small) %>%
  step_epi_ahead(num_graduates_prop, ahead = 1) %>%
  step_epi_lag(num_graduates_prop, lag = 0:2) %>%
  step_epi_naomit()
r
```

Let's apply this recipe using `prep` and `bake` to generate and view the `lag`
and `ahead` columns.

```{r view-preprocessed, include=T}
# Display a sample of the pre-processed data
bake_and_show_sample <- function(recipe, data, n = 5) {
  recipe %>%
    prep(data) %>%
    bake(new_data = data) %>%
    sample_n(n)
}

r %>% bake_and_show_sample(employ_small)
```

We can see that the `prep` and `bake` steps created new columns according to
our `epi_recipe`:

- `ahead_1_num_graduates_prop` corresponds to $y_{t+1,ijk}$
- `lag_0_num_graduates_prop`, `lag_1_num_graduates_prop`, and
`lag_2_num_graduates_prop` correspond to $y_{tijk}$, $y_{t-1,ijk}$, and $y_{t-2,ijk}$
respectively.

## Model fitting and prediction

Since our goal for now is to fit a simple autoregressive model, we can use
[`parsnip::linear_reg()`](
  https://parsnip.tidymodels.org/reference/linear_reg.html) with the default
engine `lm`, which fits a linear regression using ordinary least squares.

We will use `epi_workflow` with the `epi_recipe` we defined in the
pre-processing section along with the `parsnip::linear_reg()` model. Note
that `epi_workflow` is a container and doesn't actually do the fitting. We have
to pass the workflow into `fit()` to get our estimated model coefficients
$\widehat{\alpha}_i,\ i=0,...,3$.

```{r linearreg-wf, include=T}
wf_linreg <- epi_workflow(r, linear_reg()) %>%
  fit(employ_small)
summary(extract_fit_engine(wf_linreg))
```

This output tells us the coefficients of the fitted model; for instance,
the estimated intercept is $\widehat{\alpha}_0 =$
`r round(coef(extract_fit_engine(wf_linreg))[1], 3)` and the coefficient for
$y_{tijk}$ is
$\widehat\alpha_1 =$ `r round(coef(extract_fit_engine(wf_linreg))[2], 3)`.
The summary also tells us that all estimated coefficients are significantly
different from zero. Extracting the 95% confidence intervals for the
coefficients also leads us to
the same conclusion: all the coefficient estimates are significantly different
from 0.

```{r}
confint(extract_fit_engine(wf_linreg))
```

Now that we have our workflow, we can generate predictions from a subset of our
data. For this demo, we will predict the number of graduates using the last 2
years of our dataset.

```{r linearreg-predict, include=T}
latest <- get_predict_data(recipe = r, x = employ_small)
preds <- stats::predict(wf_linreg, latest) %>% filter(!is.na(.pred))
# Display a sample of the prediction values, excluding NAs
preds %>% sample_n(5)
```

We can do this using the `augment` function too. Note that `predict` and
`augment` both still return an `epiprocess::epi_df` with all of the keys that
were present in the original dataset.

```{r linearreg-augment}
augment(wf_linreg, latest) %>% sample_n(5)
```

## Model diagnostics

First, we'll plot the residuals (that is, $y_{tijk} - \widehat{y}_{tijk}$)
against the fitted values ($\widehat{y}_{tijk}$).

```{r lienarreg-resid-plot, include=T, fig.height = 5, fig.width = 5}
par(mfrow = c(2, 2), mar = c(5, 3, 1.2, 0))
plot(extract_fit_engine(wf_linreg))
```

The fitted values vs. residuals plot shows us that the residuals are mostly
clustered around zero, but do not form an even band around the zero line,
indicating that the variance of the residuals is not constant. Additionally,
the fitted values vs. square root of standardized residuals makes this more
obvious - the spread of the square root of standardized residuals varies with
the fitted values.

The Q-Q plot shows us that the residuals have heavier tails than a Normal
distribution. So the normality of residuals assumption doesn't hold either.

Finally, the residuals vs. leverage plot shows us that we have a few influential
points based on the Cook's distance (those outside the red dotted line).

Since we appear to be violating the linear model assumptions, we might consider
transforming our data differently, or considering a non-linear model, or
something else.

# AR model with exogenous inputs

Now suppose we want to model the 1-step-ahead 5-year employment income using
current and two previous values, while
also incorporating information from the other two time-series in our dataset:
the 2-year employment income and the number of graduates in the previous 2
years. We would do this using an autoregressive model with exogenous inputs,
defined as follows:

\[
\begin{aligned}
  y_{t+1,ijk} &=
  \alpha_0 + \alpha_1 y_{tijk} + \alpha_2 y_{t-1,ijk} + \alpha_3 y_{t-2,ijk}\\
  &\quad + \beta_1 x_{tijk} + \beta_2 x_{t-1,ijk}\\
  &\quad + \gamma_2 z_{tijk} + \gamma_2 z_{t-1,ijk} + \epsilon_{tijk}
\end{aligned}
\]

where $y_{tijk}$ is the 5-year median income (proportion) at time $t$ (in
location $i$, age group $j$ with education quality $k$),
$x_{tijk}$ is the 2-year median income (proportion) at time $t$, and
$z_{tijk}$ is the number of graduates (proportion) at time $t$.

## Pre-processing

Again, we construct an `epi_recipe` detailing the pre-processing steps.

```{r custom-arx, include=T}
rx <- epi_recipe(employ_small) %>%
  step_epi_ahead(med_income_5y_prop, ahead = 1) %>%
  # 5-year median income has current, and two lags c(0, 1, 2)
  step_epi_lag(med_income_5y_prop, lag = 0:2) %>%
  # But the two exogenous variables have current values, and 1 lag c(0, 1)
  step_epi_lag(med_income_2y_prop, lag = c(0, 1)) %>%
  step_epi_lag(num_graduates_prop, lag = c(0, 1)) %>%
  step_epi_naomit()

bake_and_show_sample(rx, employ_small)
```

## Model fitting & post-processing

Before fitting our model and making predictions, let's add some post-processing
steps using a few [`frosting`](
  https://cmu-delphi.github.io/epipredict/reference/frosting.html) layers to do
a few things:

1. Threshold our predictions to 0. We are predicting proportions, which can't
be negative. And the transformed values back to dollars and people can't be
negative either.
1. Generate prediction intervals based on residual quantiles, allowing us to
quantify the uncertainty associated with future predicted values.
1. Convert our predictions back to income values and number of graduates,
rather than standardized proportions. We do this via the frosting layer
`layer_population_scaling()`.


```{r custom-arx-post, include=T}
# Create dataframe of the sums we used for standardizing
# Only have to include med_income_5y since that is our outcome
totals <- employ_small %>%
  group_by(geo_value, age_group, edu_qual) %>%
  summarise(med_income_5y_tot = sum(med_income_5y), .groups = "drop")

# Define post-processing steps
f <- frosting() %>%
  layer_predict() %>%
  layer_naomit(.pred) %>%
  layer_threshold(.pred, lower = 0) %>%
  # 90% prediction interval
  layer_residual_quantiles(
    symmetrize = FALSE
  ) %>%
  layer_population_scaling(
    .pred, .pred_distn,
    df = totals, df_pop_col = "med_income_5y_tot"
  )

wfx_linreg <- epi_workflow(rx, parsnip::linear_reg()) %>%
  fit(employ_small) %>%
  add_frosting(f)

summary(extract_fit_engine(wfx_linreg))
```

Based on the summary output for this model, we can examine confidence intervals
and perform hypothesis tests as usual.

Let's take a look at the predictions along with their 90% prediction intervals.

```{r}
latest <- get_predict_data(recipe = rx, x = employ_small)
predsx <- predict(wfx_linreg, latest)

# Display predictions along with prediction intervals
predsx %>%
  select(
    geo_value, time_value, edu_qual, age_group,
    .pred_scaled, .pred_distn_scaled
  ) %>%
  head() %>%
  pivot_quantiles_wider(.pred_distn_scaled)
```

# Using canned forecasters

We've seen what we can do with non-epidemiological panel data using the
recipes frame, with `epi_recipe` for pre-processing, `epi_workflow` for model
fitting, and `frosting` for post-processing.

`epipredict` also comes with canned forecasters that do all of those steps
behind the scenes for some simple models. Even though we aren't working with
epidemiological data, canned forecasters still work as expected, out of the box.
We will demonstrate this with the simple
[`flatline_forecaster`](
  https://cmu-delphi.github.io/epipredict/reference/flatline_forecaster.html)
and the direct autoregressive (AR) forecaster
[`arx_forecaster`](
  https://cmu-delphi.github.io/epipredict/reference/arx_forecaster.html).

For both illustrations, we will continue to use the `employ_small` dataset
with the transformed numeric columns that are proportions within each group
by the keys in our `epi_df`.

## Flatline forecaster

In this first example, we'll use `flatline_forecaster` to make a simple
prediction of the 2-year median income for the next year, based on one previous
time point. This model is representated algebraically as:
\[y_{t+1,ijk} = y_{tijk} + \epsilon_{tijk}\]
where $y_{tijk}$ is the 2-year median income (proportion) at time $t$.

```{r flatline, include=T, warning=F}
out_fl <- flatline_forecaster(employ_small, "med_income_2y_prop",
  args_list = flatline_args_list(ahead = 1)
)

out_fl
```

## Autoregressive forecaster with exogenous inputs

In this second example, we'll use `arx_forecaster` to make a prediction of the
5-year median income based using two lags, _and_ using two lags on two exogenous
variables: 2-year median income and number of graduates.

The canned forecaster gives us a simple way of making this forecast since it
defines the recipe, workflow, and post-processing steps behind the scenes. This
is very similar to the model we introduced in the "Autoregressive Linear Model
with Exogenous Inputs" section of this article, but where all inputs have the
same number of lags.

```{r arx-lr, include=T, warning=F}
arx_args <- arx_args_list(lags = c(0L, 1L), ahead = 1L)

out_arx_lr <- arx_forecaster(employ_small, "med_income_5y_prop",
  c("med_income_5y_prop", "med_income_2y_prop", "num_graduates_prop"),
  args_list = arx_args
)

out_arx_lr
```

Other changes to the direct AR forecaster, like changing the engine, also work
as expected. Below we use a boosted tree model instead of a linear regression.

```{r arx-rf, include=T, warning=F}
out_arx_rf <- arx_forecaster(
  employ_small, "med_income_5y_prop",
  c("med_income_5y_prop", "med_income_2y_prop", "num_graduates_prop"),
  trainer = parsnip::boost_tree(mode = "regression", trees = 20),
  args_list = arx_args
)

out_arx_rf
```

# Conclusion

While the purpose of `{epipredict}` is to allow `{tidymodels}` to operate on
epidemiology data, it can be easily adapted (both the workflows and the canned
forecasters) to work for generic panel data modelling.

