---
title: Examples of Preprocessing and Models
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of Preprocessing and Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

The `epipredict` package utilizes the `tidymodels` framework, namely 
[`recipes`](https://recipes.tidymodels.org/) for 
[dplyr](https://dplyr.tidyverse.org/)-like pipeable sequences 
of feature engineering and [`parsnip`](https://parsnip.tidymodels.org/) for a 
unified interface to a range of models. 

`epipredict` has additional customized feature engineering and preprocessing 
steps, such as `step_epi_shift()`, `step_population_scaling()`, 
`step_epi_naomit()`. They can be used along with 
steps from the `recipes` package for more feature engineering. 

In this vignette, we will illustrate some examples of how to use `epipredict`
with `recipes` and `parsnip` for different purposes of epidemiology forecasting.

The remaining vignette will be split into three sections. The first section, we 
will use a Poisson regression to predict death counts. In the second section,
we will use a linear regression to predict death rates. Last but not least, we
will create a classification model for hotspot predictions. 

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(covidcast)
library(epidatr)
library(epiprocess)
library(epipredict)
library(recipes)
library(poissonreg)
```

## Poisson Regression 

During COVID-19, Center for Disease Control and Prevention (CDC) gathered models
and forecasts to characterize the state of an outbreak and its course. They use
it to inform public health decision makers on potential consequences of 
deploying control measures.

One of the outcomes that the CDC forecasts is [death counts from COVID-19](https://www.cdc.gov/coronavirus/2019-ncov/science/forecasting/forecasting-us.html).
Although there are many state-of-the-art models, we choose to use Poisson 
regression, the textbook example for modeling count data, as an illustration
for using the `epipredict` package with other existing tidymodel packages. 

```{r poisson-reg-data}
x <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj"
) %>%
  fetch_tbl() %>%
  select(geo_value, time_value, cases = value)

y <- covidcast(
  data_source = "jhu-csse",
  signals = "deaths_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj"
) %>%
  fetch_tbl() %>%
  select(geo_value, time_value, deaths = value)

case_death_counts_subset <- x %>%
  full_join(y, by = c("geo_value", "time_value")) %>%
  as_epi_df()

```

The `case_death_counts_subset` dataset comes from the `covidcast` package,and 
contains the number of confirmed cases and deaths from June 4, 2021 to 
Dec 31, 2021 in some U.S. states. 

We wish to predict the 7-day ahead death counts with lagged cases and deaths.
Furthermore, we will let each state be a dummy variable. Using differential 
intercept coefficients, we can allow for an intercept shift between states.

The model takes the form of
\begin{aligned}
\text{log}\left( \mu_{t+7} \right) = \beta_0 + \delta_1 s_{\text{state_1}} +
\delta_2 s_{\text{state_2}} + ... +  \beta_1 \text{deaths}_{t} + \nonumber \\
\beta_2 \text{deaths}_{t-7}  + \beta_3 \text{cases}_{t} + 
\beta_4 \text{cases}_{t-7}
\end{aligned}


where $\mu_{t+7} = \mathbb{E}(y_{t+7})$, and $y_{t+7}$ is assumed to follow a 
Poisson distribution with mean $\mu_{t+7}$; $s_{\text{state}}$ are dummy 
variables for each state and take values of either 0 or 1. 

Preprocessing steps will be performed using the `recipes` functions to get the 
data ready for modeling. 

But before diving into them, it will be helpful to understand what `roles` are
in the `recipes` framework. 

> `recipes` can assign one or more roles to each column in the data. The roles 
> are not restricted to a predefined set; they can be anything. 
> For most conventional situations, they are typically “predictor” and/or 
> “outcome”. Additional roles enable targeted step operations on specific 
> variables or groups of variables.
>
> In our case, the role `predictor` is given to explanatory variable that go to the 
> left-hand side of the model. The role `outcome` is the response variable 
> that we wish to predict. `geo_value` and `time_value` are predefined roles 
> that are unique to the `epipredict` package. Since we work with `epi_df` 
> objects, all datasets should have `geo_value` and `time_value` passed through
> automatically with these two roles assigned to the appropriate columns.
> 
> `recipes` package also allows [manual alterations of roles](https://recipes.tidymodels.org/reference/roles.html) 
> in bulk. There are a couple handy functions that can be used together to help us 
> manipulate variable roles easily. 
> 
>> `update_role()` alters an existing role in the recipe or assigns an initial role 
>> to variables that do not yet have a declared role.
> 
>> `add_role()` adds an additional role to variables that already have a role in 
>> the recipe, without overwriting old roles.
> 
>> `remove_role()` eliminates a single existing role in the recipe.

Notice in the following preprocessing steps, we used `add_role()` on 
`geo_value_factor` since currently the default role for it is `raw` and 
we would like the dummified variables to automatically be of role `predictor`. 

```{r}
case_death_counts_subset <- case_death_counts_subset %>%
                            mutate(geo_value_factor = as.factor(geo_value)) %>%
                            as_epi_df()

epi_recipe(case_death_counts_subset) 
               
r <- epi_recipe(case_death_counts_subset) %>%
      add_role(geo_value_factor, new_role = "predictor") %>%
      recipes::step_dummy(geo_value_factor) %>%
      step_mutate(cases = case_when(cases < 0 ~ 0, TRUE ~ cases)) %>%  
      ## NOTE: why are there negative numbers in here?
      step_mutate(deaths = case_when(deaths < 0 ~ 0, TRUE ~ deaths)) %>%
      step_epi_lag(cases, lag = c(0, 7)) %>%
      step_epi_lag(deaths, lag = c(0, 7)) %>%
      step_epi_ahead(deaths, ahead = 7, role = "outcome") %>%
      step_epi_naomit()       
```

After specifying the preprocessing steps, we will use the `parsnip` package for
modeling and getting the prediction for the 7-day ahead death counts of the 
latest available date in the dataset. 

```{r}
latest <- get_test_data(recipe = r, x = case_death_counts_subset)

wf <- epi_workflow(r, parsnip::poisson_reg()) %>% 
      fit(case_death_counts_subset)

predict(wf, latest) %>% filter(!is.na(.pred))
```

Let's take a look at the fit:
```{r, echo =FALSE}
wf$fit$fit
```

Up to now, we've used the Poisson regression to model count data. Poisson 
regression can also be used to model rate data, such as case rates or death
rates, by incorporating offset terms in the model. 

To model death rates, the Poisson regression would be expressed as:
\begin{aligned}
\text{log}\left( \mu_{t+7} \right) = \text{log(population)} + 
\beta_0 + \delta_1 s_{\text{state_1}} +
\delta_2 s_{\text{state_2}} + ... +  \beta_1 \text{deaths}_{t} + \nonumber \\
\beta_2 \text{deaths}_{t-7}  + \beta_3 \text{cases}_{t} + 
\beta_4 \text{cases}_{t-7}
\end{aligned}
where $\text{log(population)}$ is the log of the state population that was 
used to scale the count data on the left-hand side of the equation.

There are several ways to model rate data given count and population data. 
First, in the `parsnip` framework, we can specify the formula in `fit()`. 
However, by doing so we lose the ability to use the `recipes` framework to 
create new variables since new variables that do not exist in the 
original dataset such as lag and leads cannot be called directly in `fit()`. 

Alternatively, `step_population_scaling()` in the `epipredict` package can 
perform the population scaling if we provide the population data, and we can 
model this using other models such as linear regression, which we will 
illustrate in the next section.


## Linear Regression 
We wish to predict the 7-day ahead death counts with lagged cases and deaths, 
along with extra behaviour indicator inputs. Namely, we will use survey data 
from [COVID-19 Trends and Impact Survey](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html#behavior-indicators).

The survey data shows the estimated percentage of people who wore a mask for 
most or all of the time while in public in the past 7 days and the estimated 
percentage of respondents who reported that all or most people they enountered 
in public in the past 7 days maintained a distance of at least 6 feet. We will 
convert these percentages into indicator variables as follows:

```{r}
behav_ind_mask <- covidcast(data_source = "fb-survey",
                        signals = "smoothed_wwearing_mask_7d",
                        time_type = "day",
                        geo_type = "state",
                        time_values = epirange(20210604, 20211231),
                        geo_values = "ca,fl,tx,ny,nj"
                        )  %>%
            fetch_tbl() %>%
            select(geo_value, time_value, percentage = value) %>%
            mutate(mask_indicator = case_when(percentage < 50 ~ 0, 
                                              TRUE ~  1)) %>%
            select(geo_value, time_value, mask_indicator)  

behav_ind_distancing <- covidcast(data_source = "fb-survey",
                        signals = "smoothed_wothers_distanced_public",
                        time_type = "day",
                        geo_type = "state",
                        time_values = epirange(20210604, 20211231),
                        geo_values = "ca,fl,tx,ny,nj"
                        )  %>%
            fetch_tbl() %>%
            select(geo_value, time_value, percentage = value) %>%
            mutate(distancing_indicator = case_when(percentage < 20 ~ 0, 
                                              TRUE ~  1)) %>%
            select(geo_value, time_value, distancing_indicator) 
behav_ind <- behav_ind_mask %>%
  full_join(behav_ind_distancing, by = c("geo_value", "time_value")) %>%
  as_epi_df()                                
```

## Classification

## Reference
* McDonald, Daniel J., et al. "Can auxiliary indicators improve COVID-19 
forecasting and hotspot prediction?." Proceedings of the National Academy of 
Sciences 118.51 (2021): e2111453118.

## Attribution

This object contains a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API.](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html)

This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University 
on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.
