---
title: Examples of Preprocessing and Models
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of Preprocessing and Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

## Introduction

The `epipredict` package utilizes the `tidymodels` framework, namely
[`{recipes}`](https://recipes.tidymodels.org/) for
[dplyr](https://dplyr.tidyverse.org/)-like pipeable sequences
of feature engineering and [`{parsnip}`](https://parsnip.tidymodels.org/) for a
unified interface to a range of models.

`epipredict` has additional customized feature engineering and preprocessing
steps, such as `step_epi_lag()`, `step_population_scaling()`,
`step_epi_naomit()`. They can be used along with
steps from the `{recipes}` package for more feature engineering.

In this vignette, we will illustrate some examples of how to use `epipredict`
with `recipes` and `parsnip` for different purposes of epidemiological forecasting.
We will focus on basic autoregressive models, in which COVID cases and
deaths in the near future are predicted using a linear combination of cases and
deaths in the near past.

The remaining vignette will be split into three sections. The first section, we
will use a Poisson regression to predict death counts. In the second section,
we will use a linear regression to predict death rates. Last but not least, we
will create a classification model for hotspot predictions.

```{r, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(epidatr)
library(epipredict)
library(recipes)
library(workflows)
library(poissonreg)
```

## Poisson Regression

During COVID-19, the US Center for Disease Control and Prevention (CDC) collected
models
and forecasts to characterize the state of an outbreak and its course. They use
it to inform public health decision makers on potential consequences of
deploying control measures.

One of the outcomes that the CDC forecasts is [death counts from COVID-19](https://www.cdc.gov/coronavirus/2019-ncov/science/forecasting/forecasting-us.html).
Although there are many state-of-the-art models, we choose to use Poisson
regression, the textbook example for modeling count data, as an illustration
for using the `epipredict` package with other existing tidymodels packages.

```{r poisson-reg-data}
x <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj"
) %>%
  select(geo_value, time_value, cases = value)

y <- pub_covidcast(
  source = "jhu-csse",
  signals = "deaths_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj"
) %>%
  select(geo_value, time_value, deaths = value)

counts_subset <- full_join(x, y, by = c("geo_value", "time_value")) %>%
  as_epi_df()
```

The `counts_subset` dataset comes from the `epidatr` package, and
contains the number of confirmed cases and deaths from June 4, 2021 to
Dec 31, 2021 in some U.S. states.

We wish to predict the 7-day ahead death counts with lagged cases and deaths.
Furthermore, we will let each state be a dummy variable. Using differential
intercept coefficients, we can allow for an intercept shift between states.

The model takes the form
\begin{aligned}
\log\left( \mu_{t+7} \right) &= \beta_0 + \delta_1 s_{\text{state}_1} +
\delta_2 s_{\text{state}_2} + \cdots + \nonumber \\ 
&\quad\beta_1 \text{deaths}_{t} +
\beta_2 \text{deaths}_{t-7} + \beta_3 \text{cases}_{t} +
\beta_4 \text{cases}_{t-7},
\end{aligned}
where $\mu_{t+7} = \mathbb{E}(y_{t+7})$, and $y_{t+7}$ is assumed to follow a
Poisson distribution with mean $\mu_{t+7}$; $s_{\text{state}}$ are dummy
variables for each state and take values of either 0 or 1.

Preprocessing steps will be performed to prepare the
data for model fitting. But before diving into them, it will be helpful to understand what `roles` are in the `recipes` framework.

---

#### Aside on `recipes`

`recipes` can assign one or more roles to each column in the data. The roles
are not restricted to a predefined set; they can be anything.
For most conventional situations, they are typically “predictor” and/or
"outcome". Additional roles enable targeted `step_*()` operations on specific
variables or groups of variables.

In our case, the role `predictor` is given to explanatory variables on the
right-hand side of the model (in the equation above).
The role `outcome` is the response variable
that we wish to predict. `geo_value` and `time_value` are predefined roles
that are unique to the `epipredict` package. Since we work with `epi_df`
objects, all datasets should have `geo_value` and `time_value` passed through
automatically with these two roles assigned to the appropriate columns in the data.

The `recipes` package also allows [manual alterations of roles](https://recipes.tidymodels.org/reference/roles.html)
in bulk. There are a few handy functions that can be used together to help us
manipulate variable roles easily.

> `update_role()` alters an existing role in the recipe or assigns an initial role
> to variables that do not yet have a declared role.
>
> `add_role()` adds an additional role to variables that already have a role in
> the recipe, without overwriting old roles.
>
> `remove_role()` eliminates a single existing role in the recipe.

#### End aside

---

Notice in the following preprocessing steps, we used `add_role()` on
`geo_value_factor` since, currently, the default role for it is `raw`, but
we would like to reuse this variable as `predictor`s.

```{r}
counts_subset <- counts_subset %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  as_epi_df()

epi_recipe(counts_subset)

r <- epi_recipe(counts_subset) %>%
  add_role(geo_value_factor, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  ## Occasionally, data reporting errors / corrections result in negative
  ## cases / deaths
  step_mutate(cases = pmax(cases, 0), deaths = pmax(deaths, 0)) %>%
  step_epi_lag(cases, deaths, lag = c(0, 7)) %>%
  step_epi_ahead(deaths, ahead = 7, role = "outcome") %>%
  step_epi_naomit()
```

After specifying the preprocessing steps, we will use the `parsnip` package for
modeling and producing the prediction for death count, 7 days after the
latest available date in the dataset.

```{r}
latest <- get_test_data(r, counts_subset)

wf <- epi_workflow(r, parsnip::poisson_reg()) %>%
  fit(counts_subset)

predict(wf, latest) %>% filter(!is.na(.pred))
```

Note that the `time_value` corresponds to the last available date in the
training set, **NOT** to the target date of the forecast
(`r max(latest$time_value) + 7`).

Let's take a look at the fit:

```{r}
extract_fit_engine(wf)
```

Up to now, we've used the Poisson regression to model count data. Poisson
regression can also be used to model rate data, such as case rates or death
rates, by incorporating offset terms in the model.

To model death rates, the Poisson regression would be expressed as:
\begin{aligned}
\log\left( \mu_{t+7} \right) &= \log(\text{population}) +
\beta_0 + \delta_1 s_{\text{state}_1} +
\delta_2 s_{\text{state}_2} + \cdots + \nonumber \\ 
&\quad\beta_1 \text{deaths}_{t} +
\beta_2 \text{deaths}_{t-7} + \beta_3 \text{cases}_{t} +
\beta_4 \text{cases}_{t-7}
\end{aligned}
where $\log(\text{population})$ is the log of the state population that was
used to scale the count data on the left-hand side of the equation. This offset
is simply a predictor with coefficient fixed at 1 rather than estimated.

There are several ways to model rate data given count and population data.
First, in the `parsnip` framework, we could specify the formula in `fit()`.
However, by doing so we lose the ability to use the `recipes` framework to
create new variables since variables that do not exist in the
original dataset (such as, here, the lags and leads) cannot be called directly in `fit()`.

Alternatively, `step_population_scaling()` and `layer_population_scaling()`
in the `epipredict` package can perform the population scaling if we provide the
population data, which we will illustrate in the next section.

## Linear Regression

For COVID-19, the CDC required submission of case and death count predictions.
However, the Delphi Group preferred to train on rate data instead, because it
puts different locations on a similar scale (eliminating the need for location-specific intercepts).
We can use a liner regression to predict the death
rates and use state population data to scale the rates to counts.[^pois] We will do so
using `layer_population_scaling()` from the `epipredict` package.

[^pois]: We could continue with the Poisson model, but we'll switch to the Gaussian likelihood just for simplicity.

Additionally, when forecasts are submitted, prediction intervals should be
provided along with the point estimates. This can be obtained via postprocessing
using
`layer_residual_quantiles()`. It is worth pointing out, however, that
`layer_residual_quantiles()` should be used before population scaling or else
the transformation will make the results uninterpretable.

We wish, now, to predict the 7-day ahead death counts with lagged case rates and death
rates, along with some extra behaviourial predictors. Namely, we will use survey data
from [COVID-19 Trends and Impact Survey](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html#behavior-indicators).

The survey data provides the estimated percentage of people who wore a mask for
most or all of the time while in public in the past 7 days and the estimated
percentage of respondents who reported that all or most people they encountered
in public in the past 7 days maintained a distance of at least 6 feet.

State-wise population data from the 2019 U.S. Census is included in this package
and will be used in `layer_population_scaling()`.

```{r}
behav_ind_mask <- pub_covidcast(
  source = "fb-survey",
  signals = "smoothed_wwearing_mask_7d",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj"
) %>%
  select(geo_value, time_value, masking = value)

behav_ind_distancing <- pub_covidcast(
  source = "fb-survey",
  signals = "smoothed_wothers_distanced_public",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj"
) %>%
  select(geo_value, time_value, distancing = value)

pop_dat <- state_census %>% select(abbr, pop)

behav_ind <- behav_ind_mask %>%
  full_join(behav_ind_distancing, by = c("geo_value", "time_value"))
```

Rather than using raw mask-wearing / social-distancing metrics, for the sake
of illustration, we'll convert both into categorical predictors.

```{r, echo=FALSE, message=FALSE,fig.align='center', fig.width=6, fig.height=4}
library(ggplot2)
behav_ind %>%
  pivot_longer(masking:distancing) %>%
  ggplot(aes(value, fill = geo_value)) +
  geom_density(alpha = 0.5) +
  scale_fill_brewer(palette = "Set1", name = "") +
  theme_bw() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(c(0, .05))) +
  facet_wrap(~name, scales = "free") +
  theme(legend.position = "bottom")
```

We will take a subset of death rate and case rate data from the built-in dataset
`case_death_rate_subset`.

```{r}
jhu <- filter(
  case_death_rate_subset,
  time_value >= "2021-06-04",
  time_value <= "2021-12-31",
  geo_value %in% c("ca", "fl", "tx", "ny", "nj")
)
```

Preprocessing steps will again rely on functions from the `epipredict` package as well
as the `recipes` package.
There are also many functions in the `recipes` package that allow for
[scalar transformations](https://recipes.tidymodels.org/reference/#step-functions-individual-transformations),
such as log transformations and data centering. In our case, we will
center the numerical predictors to allow for a more meaningful interpretation of the
intercept.

```{r}
jhu <- jhu %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  left_join(behav_ind, by = c("geo_value", "time_value")) %>%
  as_epi_df()

r <- epi_recipe(jhu) %>%
  add_role(geo_value_factor, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  step_epi_lag(case_rate, death_rate, lag = c(0, 7, 14)) %>%
  step_mutate(
    masking = cut_number(masking, 5),
    distancing = cut_number(distancing, 5)
  ) %>%
  step_epi_ahead(death_rate, ahead = 7, role = "outcome") %>%
  step_center(contains("lag"), role = "predictor") %>%
  step_epi_naomit()
```

As a sanity check we can examine the structure of the training data:

```{r, warning = FALSE}
glimpse(slice_sample(bake(prep(r, jhu), jhu), n = 6))
```

Before directly predicting the results, we need to add postprocessing layers to
obtain the death counts instead of death rates. Note that the rates used so
far are "per 100K people" rather than "per person". We'll also use quantile
regression with the `quantile_reg` engine rather than ordinary least squares
to create median predictions and a 90% prediction interval.

```{r, warning=FALSE}
f <- frosting() %>%
  layer_predict() %>%
  layer_add_target_date("2022-01-07") %>%
  layer_threshold(.pred, lower = 0) %>%
  layer_quantile_distn() %>%
  layer_naomit(.pred) %>%
  layer_population_scaling(
    .pred, .pred_distn,
    df = pop_dat,
    rate_rescaling = 1e5,
    by = c("geo_value" = "abbr"),
    df_pop_col = "pop"
  )

wf <- epi_workflow(r, quantile_reg(quantile_levels = c(.05, .5, .95))) %>%
  fit(jhu) %>%
  add_frosting(f)

latest <- get_test_data(recipe = r, x = jhu)
p <- predict(wf, latest)
p
```

The columns marked `*_scaled` have been rescaled to the correct units, in this
case `deaths` rather than deaths per 100K people (these remain in `.pred`).

To look at the prediction intervals:

```{r}
p %>%
  select(geo_value, target_date, .pred_scaled, .pred_distn_scaled) %>%
  pivot_quantiles_wider(.pred_distn_scaled)
```

Last but not least, let's take a look at the regression fit and check the
coefficients:

```{r, echo =FALSE}
extract_fit_engine(wf)
```

## Classification

Sometimes it is preferable to create a predictive model for surges or upswings
rather than for raw values. In this case,
the target is to predict if the future will have increased case rates (denoted `up`),
decreased case rates (`down`), or flat case rates (`flat`) relative to the current
level. Such models may be
referred to as "hotspot prediction models". We will follow the analysis
in [McDonald, Bien, Green, Hu, et al.](#references) but extend the application
to predict three categories instead of two.

Hotspot prediction uses a categorical outcome variable defined in terms of the
relative change of $Y_{\ell, t+a}$ compared to $Y_{\ell, t}$.
Where $Y_{\ell, t}$ denotes the case rates in location $\ell$ at time $t$.
We define the response variables as follows:

$$
 Z_{\ell, t}=
    \begin{cases}
      \text{up}, & \text{if}\ Y^{\Delta}_{\ell, t} > 0.25 \\
      \text{down}, & \text{if}\  Y^{\Delta}_{\ell, t} < -0.20\\
      \text{flat}, & \text{otherwise}
    \end{cases}
$$

where $Y^{\Delta}_{\ell, t} = (Y_{\ell, t}- Y_{\ell, t-7})\ /\ (Y_{\ell, t-7})$.
We say location $\ell$ is a hotspot at time $t$ when $Z_{\ell,t}$ is
`up`, meaning the number of newly reported cases over the past 7 days has
increased by at least 25% compared to the preceding week. When $Z_{\ell,t}$
is categorized as `down`, it suggests that there has been at least a 20%
decrease in newly reported cases over the past 7 days (a 20% decrease is the inverse of a 25% increase). Otherwise, we will
consider the trend to be `flat`.

The expression of the multinomial regression we will use is as follows:

$$
\pi_{j}(x) = \text{Pr}(Z_{\ell,t} = j|x) = \frac{e^{g_j(x)}}{1 + \sum_{k=1}^{2}e^{g_k(x)} }
$$

where $j$ is either down, flat, or up

\begin{aligned}
g_{\text{down}}(x) &= 0.\\
g_{\text{flat}}(x) &= \log\left(\frac{Pr(Z_{\ell,t}=\text{flat}\mid x)}{Pr(Z_{\ell,t}=\text{down}\mid x)}\right) =
\beta_{10} + \beta_{11} t + \delta_{10} s_{\text{state_1}} +
\delta_{11} s_{\text{state_2}} + \cdots \nonumber \\
&\quad + \beta_{12} Y^{\Delta}_{\ell, t} +
\beta_{13} Y^{\Delta}_{\ell, t-7} \\
g_{\text{flat}}(x) &= \log\left(\frac{Pr(Z_{\ell,t}=\text{up}\mid x)}{Pr(Z_{\ell,t}=\text{down} \mid x)}\right) =
\beta_{20} + \beta_{21}t + \delta_{20} s_{\text{state_1}} +
\delta_{21} s_{\text{state}\_2} + \cdots \nonumber \\
&\quad + \beta_{22} Y^{\Delta}_{\ell, t} +
\beta_{23} Y^{\Delta}\_{\ell, t-7}
\end{aligned}

Preprocessing steps are similar to the previous models with an additional step
of categorizing the response variables. Again, we will use a subset of death rate and case rate data from our built-in dataset
`case_death_rate_subset`.

```{r}
jhu <- case_death_rate_subset %>%
  dplyr::filter(
    time_value >= "2021-06-04",
    time_value <= "2021-12-31",
    geo_value %in% c("ca", "fl", "tx", "ny", "nj")
  ) %>%
  mutate(geo_value_factor = as.factor(geo_value))

r <- epi_recipe(jhu) %>%
  add_role(time_value, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  step_growth_rate(case_rate, role = "none", prefix = "gr_") %>%
  step_epi_lag(starts_with("gr_"), lag = c(0, 7, 14)) %>%
  step_epi_ahead(starts_with("gr_"), ahead = 7, role = "none") %>%
  # note recipes::step_cut() has a bug in it, or we could use that here
  step_mutate(
    response = cut(
      ahead_7_gr_7_rel_change_case_rate,
      breaks = c(-Inf, -0.2, 0.25, Inf) / 7, # division gives weekly not daily
      labels = c("down", "flat", "up")
    ),
    role = "outcome"
  ) %>%
  step_rm(has_role("none"), has_role("raw")) %>%
  step_epi_naomit()
```

We will fit the multinomial regression and examine the predictions:

```{r, warning=FALSE}
wf <- epi_workflow(r, parsnip::multinom_reg()) %>%
  fit(jhu)

latest <- get_test_data(recipe = r, x = jhu)
predict(wf, latest) %>% filter(!is.na(.pred_class))
```

We can also look at the estimated coefficients and model summary information:

```{r}
extract_fit_engine(wf)
```

One could also use a formula in `epi_recipe()` to achieve the same results as
above. However, only one of `add_formula()`, `add_recipe()`, or
`workflow_variables()` can be specified. For the purpose of demonstrating
`add_formula` rather than `add_recipe`, we will `prep` and `bake` our recipe to
return a `data.frame` that could be used for model fitting.

```{r}
b <- bake(prep(r, jhu), jhu)

epi_workflow() %>%
  add_formula(
    response ~ geo_value + time_value + lag_0_gr_7_rel_change_case_rate +
      lag_7_gr_7_rel_change_case_rate + lag_14_gr_7_rel_change_case_rate
  ) %>%
  add_model(parsnip::multinom_reg()) %>%
  fit(data = b)
```

## Benefits of Lagging and Leading in `epipredict`

The `step_epi_ahead` and `step_epi_lag` functions in the `epipredict` package
is handy for creating correct lags and leads for future predictions.

Let's start with a simple dataset and preprocessing:

```{r}
ex <- filter(
  case_death_rate_subset,
  time_value >= "2021-12-01",
  time_value <= "2021-12-31",
  geo_value == "ca"
)

dim(ex)
```

We want to predict death rates on `r max(ex$time_value) + 7`, which is 7 days ahead of the
latest available date in our dataset.

We will compare two methods of trying to create lags and leads:

```{r}
p1 <- epi_recipe(ex) %>%
  step_epi_lag(case_rate, lag = c(0, 7, 14)) %>%
  step_epi_lag(death_rate, lag = c(0, 7, 14)) %>%
  step_epi_ahead(death_rate, ahead = 7, role = "outcome") %>%
  step_epi_naomit() %>%
  prep(ex)

b1 <- bake(p1, ex)
b1


p2 <- epi_recipe(ex) %>%
  step_mutate(
    lag0case_rate = lag(case_rate, 0),
    lag7case_rate = lag(case_rate, 7),
    lag14case_rate = lag(case_rate, 14),
    lag0death_rate = lag(death_rate, 0),
    lag7death_rate = lag(death_rate, 7),
    lag14death_rate = lag(death_rate, 14),
    ahead7death_rate = lead(death_rate, 7)
  ) %>%
  step_epi_naomit() %>%
  prep(ex)

b2 <- bake(p2, ex)
b2
```

Notice the difference in number of rows `b1` and `b2` returns. This is because
the second version, the one that doesn't use `step_epi_ahead` and `step_epi_lag`,
has omitted dates compared to the one that used the `epipredict` functions.

```{r}
dates_used_in_training1 <- b1 %>%
  select(-ahead_7_death_rate) %>%
  na.omit() %>%
  pull(time_value)
dates_used_in_training1

dates_used_in_training2 <- b2 %>%
  select(-ahead7death_rate) %>%
  na.omit() %>%
  pull(time_value)
dates_used_in_training2
```

The model that is trained based on the `{recipes}` functions will predict 7 days ahead from
`r max(dates_used_in_training2)`
instead of 7 days ahead from `r max(dates_used_in_training1)`.

## References

McDonald, Bien, Green, Hu, et al. "Can auxiliary indicators improve COVID-19
forecasting and hotspot prediction?." Proceedings of the National Academy of
Sciences 118.51 (2021): e2111453118. [doi:10.1073/pnas.2111453118](https://doi.org/10.1073/pnas.2111453118)

## Attribution

This object contains a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API.](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html)

This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University
on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.
