---
title: Examples of Preprocessing and Models
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of Preprocessing and Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

The `epipredict` package utilizes the `tidymodels` framework, namely 
[`recipes`](https://recipes.tidymodels.org/) for 
[dplyr](https://dplyr.tidyverse.org/)-like pipeable sequences 
of feature engineering and [`parsnip`](https://parsnip.tidymodels.org/) for a 
unified interface to a range of models. 

`epipredict` has additional customized feature engineering and preprocessing 
steps, such as `step_epi_shift()`, `step_population_scaling()`, 
`step_epi_naomit()`. They can be used along with 
steps from the `recipes` package for more feature engineering. 

In this vignette, we will illustrate some examples of how to use `epipredict`
with `recipes` and `parsnip` for different purposes of epidemiology forecasting.
We will focus on simple basic autoregressive models, in which COVID cases and 
deaths in the near future are predicted using a linear combination of cases and 
deaths in the near past.

The remaining vignette will be split into three sections. The first section, we 
will use a Poisson regression to predict death counts. In the second section,
we will use a linear regression to predict death rates. Last but not least, we
will create a classification model for hotspot predictions. 

```{r, warning=FALSE, message=FALSE}
# library(covidcast)
library(tidyr)
library(dplyr)
library(epidatr)
library(epiprocess)
library(epipredict)
library(recipes)
library(parsnip)
library(workflows)
library(poissonreg)
```

## Poisson Regression 

During COVID-19, Center for Disease Control and Prevention (CDC) gathered models
and forecasts to characterize the state of an outbreak and its course. They use
it to inform public health decision makers on potential consequences of 
deploying control measures.

One of the outcomes that the CDC forecasts is [death counts from COVID-19](https://www.cdc.gov/coronavirus/2019-ncov/science/forecasting/forecasting-us.html).
Although there are many state-of-the-art models, we choose to use Poisson 
regression, the textbook example for modeling count data, as an illustration
for using the `epipredict` package with other existing tidymodel packages. 

```{r poisson-reg-data}
x <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj") %>%
  fetch_tbl() %>%
  select(geo_value, time_value, cases = value)

y <- covidcast(
  data_source = "jhu-csse",
  signals = "deaths_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj") %>%
  fetch_tbl() %>%
  select(geo_value, time_value, deaths = value)

case_death_counts_subset <- x %>%
  full_join(y, by = c("geo_value", "time_value")) %>%
  as_epi_df()
```

The `case_death_counts_subset` dataset comes from the `epidatr` package,and 
contains the number of confirmed cases and deaths from June 4, 2021 to 
Dec 31, 2021 in some U.S. states. 

We wish to predict the 7-day ahead death counts with lagged cases and deaths.
Furthermore, we will let each state be a dummy variable. Using differential 
intercept coefficients, we can allow for an intercept shift between states.

The model takes the form of
\begin{aligned}
\text{log}\left( \mu_{t+7} \right) = \beta_0 + \delta_1 s_{\text{state_1}} +
\delta_2 s_{\text{state_2}} + ... +  \beta_1 \text{deaths}_{t} + \nonumber \\
\beta_2 \text{deaths}_{t-7}  + \beta_3 \text{cases}_{t} + 
\beta_4 \text{cases}_{t-7}
\end{aligned}


where $\mu_{t+7} = \mathbb{E}(y_{t+7})$, and $y_{t+7}$ is assumed to follow a 
Poisson distribution with mean $\mu_{t+7}$; $s_{\text{state}}$ are dummy 
variables for each state and take values of either 0 or 1. 

Preprocessing steps will be performed using the `recipes` functions to get the 
data ready for modeling. 

But before diving into them, it will be helpful to understand what `roles` are
in the `recipes` framework. 

---

#### Aside on `recipes`

`recipes` can assign one or more roles to each column in the data. The roles 
are not restricted to a predefined set; they can be anything. 
For most conventional situations, they are typically “predictor” and/or 
"outcome". Additional roles enable targeted `step_*()` operations on specific 
variables or groups of variables.

In our case, the role `predictor` is given to explanatory variable that go to the 
left-hand side of the model. The role `outcome` is the response variable 
that we wish to predict. `geo_value` and `time_value` are predefined roles 
that are unique to the `epipredict` package. Since we work with `epi_df` 
objects, all datasets should have `geo_value` and `time_value` passed through
automatically with these two roles assigned to the appropriate columns.
 
The `recipes` package also allows [manual alterations of roles](https://recipes.tidymodels.org/reference/roles.html) 
in bulk. There are a couple handy functions that can be used together to help us 
manipulate variable roles easily. 

> `update_role()` alters an existing role in the recipe or assigns an initial role 
> to variables that do not yet have a declared role.
> 
> `add_role()` adds an additional role to variables that already have a role in 
> the recipe, without overwriting old roles.
> 
> `remove_role()` eliminates a single existing role in the recipe.

#### End aside

---

Notice in the following preprocessing steps, we used `add_role()` on 
`geo_value_factor` since currently the default role for it is `raw` and 
we would like the dummified variables to automatically be `predictor`'s. 

```{r}
case_death_counts_subset <- case_death_counts_subset %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  as_epi_df()

epi_recipe(case_death_counts_subset)

r <- epi_recipe(case_death_counts_subset) %>%
  add_role(geo_value_factor, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  ## Occasionally, data reporting errors / corrections result in negative
  ## cases / deaths
  step_mutate(cases = pmax(cases, 0), deaths = pmax(deaths, 0)) %>%  
  step_epi_lag(cases, deaths, lag = c(0, 7)) %>%
  step_epi_ahead(deaths, ahead = 7, role = "outcome") %>%
  step_epi_naomit()
```

After specifying the preprocessing steps, we will use the `parsnip` package for
modeling and getting the prediction for death count, 7 days after the
latest available date in the dataset. 

```{r}
latest <- get_test_data(r, case_death_counts_subset)

wf <- epi_workflow(r, parsnip::poisson_reg()) %>% 
      fit(case_death_counts_subset)

predict(wf, latest) %>% filter(!is.na(.pred))
```

Note that the `time_value` corresponds to the last available date in the 
training set, **NOT** to the target date of the forecast 
(`r max(latest$time_value) + 7`).


Let's take a look at the fit:
```{r}
extract_fit_engine(wf)
```

Up to now, we've used the Poisson regression to model count data. Poisson 
regression can also be used to model rate data, such as case rates or death
rates, by incorporating offset terms in the model. 

To model death rates, the Poisson regression would be expressed as:
\begin{aligned}
\text{log}\left( \mu_{t+7} \right) = \text{log(population)} + 
\beta_0 + \delta_1 s_{\text{state_1}} +
\delta_2 s_{\text{state_2}} + ... +  \beta_1 \text{deaths}_{t} + \nonumber \\
\beta_2 \text{deaths}_{t-7}  + \beta_3 \text{cases}_{t} + 
\beta_4 \text{cases}_{t-7}
\end{aligned}
where $\text{log(population)}$ is the log of the state population that was 
used to scale the count data on the left-hand side of the equation.

There are several ways to model rate data given count and population data. 
First, in the `parsnip` framework, we can specify the formula in `fit()`. 
However, by doing so we lose the ability to use the `recipes` framework to 
create new variables since new variables that do not exist in the 
original dataset such as lag and leads cannot be called directly in `fit()`. 

Alternatively, `step_population_scaling()` and `layer_population_scaling()` 
in the `epipredict` package can perform the population scaling if we provide the 
population data, and we can model this using other models such as linear 
regression, which we will illustrate in the next section.


## Linear Regression 

For COVID-19, the CDC required submission of case and death count predictions. 
However, the Delphi Group preferred to train on rate data instead, because it 
puts different locations on a similar scale. 
We can use a liner regression to predict the death
rates and use state population data to scale the rates to counts. We will do so 
using `layer_population_scaling()` from the `epipredict` package. 

Additionally, when forecasts are submitted, prediction intervals should be 
provided along with the point estimates. This can be obtained via postprocessing
`layer_residual_quantiles()`. Although worth pointing out that 
`layer_residual_quantiles()` should be used before population scaling or else 
the transformation will make the results uninterpretable. 

We wish to predict the 7-day ahead death counts with lagged case rates and death
rates, along with extra behaviour indicator inputs. Namely, we will use survey data 
from [COVID-19 Trends and Impact Survey](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html#behavior-indicators).

The survey data shows the estimated percentage of people who wore a mask for 
most or all of the time while in public in the past 7 days and the estimated 
percentage of respondents who reported that all or most people they encountered 
in public in the past 7 days maintained a distance of at least 6 feet. 

State-wise population data from the 2019 U.S. Census is included in this package
and will be used in `layer_population_scaling()`. 
```{r}
behav_ind_mask <- covidcast(
  data_source = "fb-survey",
  signals = "smoothed_wwearing_mask_7d",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj")  %>%
  fetch_tbl() %>%
  select(geo_value, time_value, masking = value)

behav_ind_distancing <- covidcast(
  data_source = "fb-survey",
  signals = "smoothed_wothers_distanced_public",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20210604, 20211231),
  geo_values = "ca,fl,tx,ny,nj")  %>%
  fetch_tbl() %>%
  select(geo_value, time_value, distancing = value) 

pop_dat <- state_census %>% select(abbr, pop)

behav_ind <- behav_ind_mask %>%
  full_join(behav_ind_distancing, by = c("geo_value", "time_value")) 
```

Rather than using raw mask-wearing / social-distancing metrics, for the sake
of illustration, we'll convert both into categorical predictors.

```{r, echo=FALSE, message=FALSE,fig.align='center', fig.width=6, fig.height=4}
library(ggplot2)
behav_ind %>%
  pivot_longer(masking:distancing) %>%
  ggplot(aes(value, fill = geo_value)) +
  geom_density(alpha = 0.5) +
  scale_fill_brewer(palette = "Set1", name = "") +
  theme_bw() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = expansion(c(0,.05))) +
  facet_wrap(~name, scales = "free") +
  theme(legend.position = "bottom")
```

We will take a subset of death rate and case rate data from the built-in dataset 
`case_death_rate_subset`.

```{r}
jhu <- case_death_rate_subset %>%
  dplyr::filter(
    time_value >= "2021-06-04", 
    time_value <= "2021-12-31",
    geo_value %in% c("ca","fl","tx","ny","nj"))
```

Preprocessing steps will rely on functions from the `epipredict` package as well
as the `recipes` package:

Additionally, there are also many functions in the `recipes` package that allow for 
[scalar transformations](https://recipes.tidymodels.org/reference/#step-functions-individual-transformations),
such as log transformations and data centering. In our case, we will 
center the numerical predictors to allow for meaningful interpretation of the 
intercept. 

```{r}
jhu <- jhu %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  left_join(behav_ind, by = c("geo_value", "time_value")) %>%
  as_epi_df()
            
r <- epi_recipe(jhu) %>%
  add_role(geo_value_factor, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  step_epi_lag(case_rate, death_rate, lag = c(0, 7, 14)) %>%
  step_mutate(masking = cut_number(masking, 5), 
              distancing = cut_number(distancing, 5)) %>%
  step_epi_ahead(death_rate, ahead = 7, role = "outcome") %>%
  step_center(contains("lag"), role = "predictor") %>%
  step_epi_naomit()
```

As a sanity check we can see what the training data will look like
```{r, warning = FALSE}
glimpse(slice_sample(bake(prep(r, jhu), jhu), n = 6))
```

Before directly predicting the results, we need to add postprocessing layers to
obtain the death counts instead of death rates. Note that the rates used so
far are "per 100K people" rather than "per person".

```{r}
f <- frosting() %>%
  layer_predict() %>%
  layer_add_target_date("2022-01-07") %>% 
  layer_threshold(.pred, lower = 0) %>%
  layer_residual_quantiles(probs = c(0.05, 0.95), symmetrize = FALSE) %>%
  layer_naomit(.pred) %>%
  layer_population_scaling(
    .pred, .pred_distn, df = pop_dat, rate_rescaling = 1e5,
    by = c("geo_value" = "abbr"), df_pop_col = "pop")

wf <- epi_workflow(r, parsnip::linear_reg()) %>%
    fit(jhu) %>%
    add_frosting(f)

latest <- get_test_data(recipe = r, x = jhu)
p <- predict(wf, latest)
p
```

The columns marked `*_scaled` have been rescaled to the correct units, in this
case `deaths` rather than deaths per 100K people (these remain in `.pred`).

To look at the prediction intervals:
```{r}
p %>%
  select(geo_value, target_date, .pred_scaled, .pred_distn_scaled) %>%
  mutate(.pred_distn_scaled = nested_quantiles(.pred_distn_scaled)) %>%
  unnest(.pred_distn_scaled) %>%
  pivot_wider(names_from = tau, values_from = q)
```


Last but not least, let's take a look at the regression fit and check the 
coefficients:
```{r, echo =FALSE}
summary(extract_fit_engine(wf))
```

## Classification

Sometimes it is preferable to create a predictive model for surges or upswings
rather than for raw values. In this case,
the target is to predict if the future will have increased case rates(up),
decreased case rates (down), or flat case rates (flat) relative to the current
level. Such model are often 
referred to as hotspot prediction models. We will refer to the model notations 
from [McDonald, Bien, Green, Hu, et al.](#references) but extend the application
to predict three categories instead of two. 

Hotspot prediction predicts a categorical variable defined in terms of the 
relative change of $Y_{\ell, t+a}$ compared to $Y_{\ell, t}$. 
Where $Y_{\ell, t}$ denotes the case rates in location $\ell$ at time $t$. 
We define the response variables as follows:

$$
 Z_{\ell, t}=
    \begin{cases}
      \text{up}, & \text{if}\ Y^{\Delta}_{\ell, t} > 0.25 \\ 
      \text{down}, & \text{if}\  Y^{\Delta}_{\ell, t} < -0.25\\
      \text{flat}, & \text{otherwise}
    \end{cases}
$$

where $Y^{\Delta}_{\ell, t} = (Y_{\ell, t}- Y_{\ell, t-7})/(Y_{\ell, t-7})$. 
We say location $\ell$ is a hotspot at time $t$ when $Z_{\ell,t}$ is categorized
as 'up', meaning the number of newly reported cases over the past 7 days has 
increased by at least 25% compared to the preceding week. When $Z_{\ell,t}$ 
is categorized as 'down', it suggests that there has been at least a 25% 
decrease in newly reported cases over the past 7 days. Otherwise, we will 
consider the trends to be flat. 

The expression of the multinomial regression we will model is as follows:

$$
\pi_{j}(x) = \text{Pr}(Y = j|x) = \frac{e^{g_j(x)}}{\sum_{k=0}^2 g_j(x) }
$$
where $j$ is either down (`Y=0`), flat (`Y=1`), or up (`Y=2`),
$$g_0(x) = 0$$

\begin{aligned}
g_1(x)= \text{ln}\left(\frac{Pr(Y=1|x)}{Pr(Y=0|x)}\right) &= 
\beta_{10} + \beta_{11}* x_{\text{time_value}} + \delta_{10} s_{\text{state_1}} +
\delta_{11} s_{\text{state_2}} + ... \nonumber \\
&\quad + \beta_{12} Y^{\Delta}_{\ell, t} +
\beta_{13} Y^{\Delta}_{\ell, t-7}
\end{aligned}

\begin{aligned}
g_2(x)= \text{ln}\left(\frac{Pr(Y=2|x)}{Pr(Y=0|x)}\right) &= 
\beta_{20} + \beta_{21}* x_{\text{time_value}} + \delta_{20} s_{\text{state_1}} +
\delta_{21} s_{\text{state_2}} + ... \nonumber \\
&\quad + \beta_{22} Y^{\Delta}_{\ell, t} +
\beta_{23} Y^{\Delta}_{\ell, t-7}
\end{aligned}



Preprocessing steps are similar to the previous models with an additional step 
of categorizing the response variables. 

We will take a subset of death rate and case rate data from our built-in dataset 
`case_death_rate_subset`.
```{r}
jhu <- case_death_rate_subset %>%
  dplyr::filter(time_value >= "2021-06-04", 
                time_value <= "2021-12-31",
  geo_value %in% c("ca","fl","tx","ny","nj")) %>%
  mutate(geo_value_factor = as.factor(geo_value)) %>%
  as_epi_df()

r <- epi_recipe(jhu) %>%
  add_role(time_value, new_role = "predictor") %>%
  step_dummy(geo_value_factor) %>%
  step_epi_lag(case_rate, lag = c(0, 7, 14)) %>%
  step_epi_ahead(case_rate, ahead = 7, role = "predictor") %>%
  step_mutate(
    pct_diff_ahead = case_when(
      lag_7_case_rate == 0 ~ 0,
      TRUE ~ (ahead_7_case_rate - lag_0_case_rate) / lag_0_case_rate),
    pct_diff_wk1 = case_when(
      lag_7_case_rate == 0 ~ 0, 
      TRUE ~ (lag_0_case_rate - lag_7_case_rate) / lag_7_case_rate),
    pct_diff_wk2 = case_when(
      lag_14_case_rate == 0 ~ 0,
      TRUE ~ (lag_7_case_rate - lag_14_case_rate) / lag_14_case_rate)) %>%
  step_mutate(
    response = case_when(
      pct_diff_ahead < -0.20 ~ "down",
      pct_diff_ahead > 0.25 ~ "up",
      TRUE ~ "flat"), 
    role = "outcome") %>% 
  step_rm(death_rate, case_rate, lag_0_case_rate,  lag_7_case_rate, 
          lag_14_case_rate, ahead_7_case_rate, pct_diff_ahead) %>%
  step_epi_naomit()
```

We will fit the multinomial regression and take a look at the predictions:

```{r, warning=FALSE}
wf <- epi_workflow(r, parsnip::multinom_reg()) %>%
    fit(jhu)

latest <- get_test_data(recipe = r, x = jhu)
predict(wf, latest) %>% filter(!is.na(.pred_class))
```

We can also take a look at the fit:
```{r}
extract_fit_engine(wf)
```

One could also use a formula in `epi_recipe()` to achieve the same results as 
above. However, only one of `add_formula()`, `add_recipe()`, or 
`workflow_variables()` must be specified. For the purpose of demonstrating 
`add_formula` rather than `add_recipe`, we will `prep` and `bake` our recipe to
return a `data.frame` that could be used for model fitting.
```{r}
b <- bake(prep(r, jhu), jhu)

epi_workflow() %>%
  add_formula(response ~ geo_value + pct_diff_wk1 + pct_diff_wk2) %>%
  add_model(parsnip::multinom_reg()) %>%
  fit(data = b)
```

## Benefits of Lagging and Leading in `epipredict`

The `step_epi_ahead` and `step_epi_lag` functions in the `epipredict` package
is handy for creating correct lags and leads for future predictions. 

Let's start with a simple dataset and preprocessing:
```{r}
ex <- case_death_rate_subset %>%
  dplyr::filter(time_value >= "2021-12-01", 
                time_value <= "2021-12-31",
                geo_value == "ca")

dim(ex)
```

We want to predict death rates on `2022-01-07`, which is 7 days ahead from the 
latest available date in our dataset. 

We will compare two methods of trying to create lags and leads:
```{r}
p1 <- epi_recipe(ex) %>%
  step_epi_lag(case_rate, lag = c(0, 7, 14)) %>%
  step_epi_lag(death_rate, lag = c(0, 7, 14)) %>%
  step_epi_ahead(death_rate, ahead = 7, role = "outcome") %>%
  step_epi_naomit() %>%
  prep()

b1 <- bake(p1, ex)
b1


p2 <- epi_recipe(ex) %>%
  step_mutate(lag0case_rate = lag(case_rate, 0),
              lag7case_rate = lag(case_rate, 7),
              lag14case_rate = lag(case_rate, 14),
              lag0death_rate = lag(death_rate, 0),
              lag7death_rate = lag(death_rate, 7),
              lag14death_rate = lag(death_rate, 14),
              ahead7death_rate = lead(death_rate, 7)) %>%
  step_epi_naomit() %>%
  prep()

b2 <- bake(p2, ex)
b2
```

Notice the difference in number of rows `b1` and `b2` returns. This is because 
the second version, the one that doesn't use `step_epi_ahead` and `step_epi_lag`,
has omitted dates compared to the one that used the `epipredict` functions.
```{r}
dates_used_in_training1 <-
 b1 %>% select(- ahead_7_death_rate) %>% na.omit() %>% select(time_value)

dates_used_in_training1

dates_used_in_training2 <- 
  b2 %>% select(- ahead7death_rate) %>% na.omit() %>% select(time_value)

dates_used_in_training2
```

The model that is trained without the functions is predicting 7 days ahead from 
`r max(dates_used_in_training2$time_value)`
instead of 7 days ahead from `r max(dates_used_in_training1$time_value)`.

## References

McDonald, Bien, Green, Hu, et al. "Can auxiliary indicators improve COVID-19 
forecasting and hotspot prediction?." Proceedings of the National Academy of 
Sciences 118.51 (2021): e2111453118. [doi:10.1073/pnas.2111453118](
https://doi.org/10.1073/pnas.2111453118)

## Attribution

This object contains a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API.](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html)

This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University 
on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.
