---
title: "Visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Good visualizations help the user understand what is going on. One may use
`plot.covidcast_signal` from the `covidcast` package to plot US maps, including
county maps. However, `covidcast` will not support maps unrelated to the US.
One may use `plot` and `ggplot` for various plots
including scatterplots. Finally, users can add animation and interactive
features on the plot.

## covidcast_signal maps

`epipredict` supports visualizations features that are a part of COVIDcast.
However, it is not intended for
those who use third-party data visualization software such as Power BI and
Tableau. Despite that, visualizations go alongside cmu-delphi's covidcast
library.

For the first example, we are tracking the cumulative number of COVID cases
by state up until 2021-12-31.

```{r}
library(covidcast)
library(dplyr)
x <- case_death_rate_subset %>%
  filter(!(geo_value %in% c("as","cm","gu","mp","vi")),
         time_value <= "2021-12-31") %>%
  group_by(geo_value) %>%
  summarize(value = sum(case_rate) / 1000) %>%
  mutate(time_value = as.Date("2021-12-31"))

cc <- as.covidcast_signal(x,
                          geo_type = "state",
                          data_source = "covid-tracking",
                          signal = "cumulative_case_rate_thousands")

plot(cc)
```

We need not use a dataset that's already stored on epipredict or epiprocess.

Keep in mind that only 'county', 'state', 'hrr' and 'msa' are supported
for map plots; 'nation', for example, is NOT supported for a covidcast_signal 
plot. Thus, one would need to load separate packages, etc., do a plot of
something such as a world map or a map of Canada (as not even "province"
is an option for covidcast_signal's geo_type).

As a matter of fact, it's possible to use data unrelated to COVID,
such as data relating to the 2020 elections. As always, one should follow
guidelines relating to good data visualizations, such as by making colours
match with blue for Democrat and red for Republican.

This chart shows the percentage of people under the age of 18 in each state.
This map includes all 50 states, Washington D.C. and Puerto Rico.

```{r}
library(readr)
library(tidyr)

y <- read_csv("https://raw.githubusercontent.com/jakevdp/data-USstates/master/state-population.csv")

y2 <- y %>%
  rename(geo_value = 1, time_value = year) %>%
  spread(ages,population) %>%
  mutate(value = under18 / total * 100, geo_value = tolower(geo_value)) %>%
  select(-total,-under18) %>%
  filter(time_value >= 2009)

cc2 <- as.covidcast_signal(y2,
                          geo_type = "state",
                          data_source = "census",
                          signal = "percent_under_18")

plot(cc2,choro_col = c("#C040C0", "#C04040", "#C08040", "#C0C040",
                       "#40C040", "#40C0C0", "#4040C0"))
```

## Non-USA plots

If a user wants to map something such as Canadian provinces or a world map,
they will need to install a separate package and use features outside of
covidcast_signal, as those are not recognized geo_type's.

There may be multiple ways to achieve this, whether it is downloading a package
or drawing areas with a package that supports ArcGIS.

Note that both epipredict's epi_df and epi_archive both recognize custom
geo_type's. However, this does not mean they enable different types of maps
without downloading other packages.

## Base R plotting

Base R has its own plotting with `plot`. However, we will not cover this here
as `ggplot` arguably looks nicer and is a common feature to learn about given
that it is a part of the well-used `tidyverse` package.

## Other plots using ggplot2

`ggplot2` is a part of `tidyverse` that enables users to do plots outside of
base R's plots that may not look as nice.

A plot of the USA is just one plot; there are other plots, such as line plots,
bar plots, etc. This is helpful if we want to plot something that is not
recognized under `plot.covidcast_signal`, such as a plot of European countries.
(However, it may be better to just use a different map feature for plotting
other countries, etc., even if it is more complicated.)

For this, we are looking to observe COVID-related data around the world.

```{r}
library(ggplot2)

z <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv")
```

We don't even need to look at COVID-related data. Here is an example of such,
finding the relationship between Human Development Index and the logarithm of
the number of hospital beds per thousand people on 2022-06-01.

```{r}
z2 <- z %>%
  drop_na(continent) %>%
  select(location,date,hospital_beds_per_thousand,human_development_index) %>%
  filter(date == "2022-06-01") 

ggplot(z2,aes(x=human_development_index,y=log(hospital_beds_per_thousand))) +
  geom_point()
```
As we can see, there is moderate
positive correlation between the how developed a country is and the number
of hospital beds a country has.

We can grab COVID-related data, such as with vaccination rates of different
countries.

```{r}
z3 <- z %>%
  drop_na(continent) %>%
  select(location,date,total_cases_per_million,total_vaccinations_per_hundred,
         human_development_index)  %>%
  filter(date == "2021-07-01") %>%
  drop_na(total_vaccinations_per_hundred) %>%
  mutate(log_cases_pm = log(total_cases_per_million),
         log_vacs_pm = log(10000 * total_vaccinations_per_hundred))

ggplot(z3,aes(x=human_development_index,y=log_vacs_pm,colour=log_cases_pm)) +
  geom_point() +
  scale_colour_viridis_c()

z4 <- z3 %>%
  drop_na() %>%
  filter(total_cases_per_million > 0, total_vaccinations_per_hundred > 0)

lm1 <- lm(log_vacs_pm ~ human_development_index,z4)

qplot(lm1$fitted.values,lm1$residuals) +
  labs(x = "Fitted values", y = "Residuals",
       title = "Fitted values vs. residuals of a linear model with interaction")

summary(lm1)
```
This chart shows that the percentage of people vaccinated as of 2022-07-01.
Countries with higher COVID cases per capita may tend to be able to afford more
tests, although we may not be able to conclude that from the limited information
we have.

Trying to find a meaningful relationship between various parameters is easier
said than done when you have to account for things such as heteroskedasticity,
collinearity and non-linear relationships. In addition, one cannot conclude that
correlation implies causality.

## Beyond COVIDcast and ggplot2

Note that many things can be done with R that are beyond the scope of the
audience, such as animation and interactive visualizations. However, one may
learn about packages that support these here:

https://r-graph-gallery.com/
