---
title: "Demonstrations of sliding AR and ARX forecasters"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

```{r pkgs}
library(epipredict)
library(epidatr)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(magrittr)
library(purrr)
```

# Demonstrations of sliding AR and ARX forecasters

A key function from the epiprocess package is `epix_slide()` (refer to the
following vignette for the basics of the function: ["Work with archive objects
and data
revisions"](https://cmu-delphi.github.io/epiprocess/articles/archive.html))
which allows performing version-aware computations. That is, the function only
uses data that would have been available as of time t for that reference time.

In this vignette, we use `epix_slide()` for backtesting our `arx_forecaster` on
historical COVID-19 case data from the US and from Canada. We first examine the
results from a version-unaware forecaster, comparing two different fitting
engines and then we contrast this with version-aware forecasting. The former
will proceed by constructing an `epi_archive` that erases its version
information and then use `epix_slide()` to forecast the future. The latter will
keep the versioned data and proceed similarly by using `epix_slide()` to
forecast the future.

## Comparing different forecasting engines

### Example using CLI and case data from US states

First, we download the version history (ie. archive) of the percentage of
doctor's visits with CLI (COVID-like illness) computed from medical insurance
claims and the number of new confirmed COVID-19 cases per 100,000 population
(daily) for all 50 states from the COVIDcast API.

<details>

<summary>Load a data archive</summary>

We process as before, with the modification that we use `sync = locf` in
`epix_merge()` so that the last version of each observation can be carried
forward to extrapolate unavailable versions for the less up-to-date input
archive.

```{r grab-epi-data}
theme_set(theme_bw())

y <- readRDS("all_states_covidcast_signals.rds") %>%
  purrr::map(~ select(.x, geo_value, time_value, version = issue, value))

x <- epix_merge(
  y[[1]] %>% rename(percent_cli = value) %>% as_epi_archive(compactify = FALSE),
  y[[2]] %>% rename(case_rate = value) %>% as_epi_archive(compactify = FALSE),
  sync = "locf",
  compactify = TRUE
)
rm(y)
```

</details>

We then obtaining the latest snapshot of the data and proceed to fake the
version information by setting `version = time_value`. This has the effect of
obtaining data that arrives exactly at the day of the time_value.

```{r arx-kweek-preliminaries, warning = FALSE}
# Latest snapshot of data, and forecast dates
x_latest <- epix_as_of(x, version = max(x$versions_end)) %>%
  mutate(version = time_value) %>%
  as_epi_archive()
fc_time_values <- seq(
  from = as.Date("2020-08-01"),
  to = as.Date("2021-11-01"),
  by = "1 month"
)
aheads <- c(7, 14, 21, 28)

forecast_k_week_ahead <- function(epi_archive, outcome, predictors, ahead = 7, engine) {
  epi_archive %>%
    epix_slide(
      .f = function(x, gk, rtv) {
        arx_forecaster(
          x, outcome, predictors, engine,
          args_list = arx_args_list(ahead = ahead)
        )$predictions %>%
          mutate(engine_type = engine$engine) %>%
          pivot_quantiles_wider(.pred_distn)
      },
      .before = 120,
      .versions = fc_time_values
    )
}
```

```{r make-arx-kweek}
# Generate the forecasts and bind them together
fc <- bind_rows(
  map(aheads, ~ forecast_k_week_ahead(
    x_latest,
    outcome = "case_rate",
    predictors = c("case_rate", "percent_cli"),
    ahead = .x,
    engine = linear_reg()
  )),
  map(aheads, ~ forecast_k_week_ahead(
    x_latest,
    outcome = "case_rate",
    predictors = c("case_rate", "percent_cli"),
    ahead = .x,
    engine = rand_forest(mode = "regression")
  ))
)
```

Here, `arx_forecaster()` does all the heavy lifting. It creates leads of the
target (respecting time stamps and locations) along with lags of the features
(here, the response and doctors visits), estimates a forecasting model using the
specified engine, creates predictions, and non-parametric confidence bands.

To see how the predictions compare, we plot them on top of the latest case
rates. Note that even though we've fitted the model on all states, we'll just
display the results for two states, California (CA) and Florida (FL), to get a
sense of the model performance while keeping the graphic simple.

<details>

<summary>Code for plotting</summary>
```{r plot-arx, message = FALSE, warning = FALSE}
fc_cafl <- fc %>%
  tibble() %>%
  filter(geo_value %in% c("ca", "fl"))
x_latest_cafl <- x_latest$DT %>%
  tibble() %>%
  filter(geo_value %in% c("ca", "fl"))

p1 <- ggplot(fc_cafl, aes(target_date, group = forecast_date, fill = engine_type)) +
  geom_line(
    data = x_latest_cafl, aes(x = time_value, y = case_rate),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`), alpha = 0.4) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_grid(vars(geo_value), vars(engine_type), scales = "free") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Date", y = "Reported COVID-19 case rates") +
  theme(legend.position = "none")
```
</details>

```{r show-plot1, fig.width = 9, fig.height = 6, echo=FALSE}
p1
```

For the two states of interest, simple linear regression clearly performs better
than random forest in terms of accuracy of the predictions and does not result
in such in overconfident predictions (overly narrow confidence bands). Though,
in general, neither approach produces amazingly accurate forecasts. This could
be because the behaviour is rather different across states and the effects of
other notable factors such as age and public health measures may be important to
account for in such forecasting. Including such factors as well as making
enhancements such as correcting for outliers are some improvements one could
make to this simple model.[^1]

[^1]: Note that, despite the above caveats, simple models like this tend to
out-perform many far more complicated models in the online Covid forecasting due
to those models high variance predictions.


### Example using case data from Canada

<details>

<summary>Data and forecasts. Similar to the above.</summary>

By leveraging the flexibility of `epiprocess`, we can apply the same techniques
to data from other sources. Since some collaborators are in British Columbia,
Canada, we'll do essentially the same thing for Canada as we did above.

The [COVID-19 Canada Open Data Working Group](https://opencovid.ca/) collects
daily time series data on COVID-19 cases, deaths, recoveries, testing and
vaccinations at the health region and province levels. Data are collected from
publicly available sources such as government datasets and news releases.
Unfortunately, there is no simple versioned source, so we have created our own
from the Github commit history.

First, we load versioned case rates at the provincial level. After converting
these to 7-day averages (due to highly variable provincial reporting
mismatches), we then convert the data to an `epi_archive` object, and extract
the latest version from it. Finally, we run the same forcasting exercise as for
the American data, but here we compare the forecasts produced from using simple
linear regression with those from using boosted regression trees.

```{r get-can-fc, warning = FALSE}
# source("drafts/canada-case-rates.R)
can <- readRDS(system.file(
  "extdata", "can_prov_cases.rds",
  package = "epipredict", mustWork = TRUE
)) %>%
  group_by(version, geo_value) %>%
  arrange(time_value) %>%
  mutate(cr_7dav = RcppRoll::roll_meanr(case_rate, n = 7L)) %>%
  as_epi_archive(compactify = TRUE)

can_latest <- epix_as_of(can, version = max(can$DT$version)) %>%
  mutate(version = time_value) %>%
  as_epi_archive()

# Generate the forecasts, and bind them together
can_fc <- bind_rows(
  map(
    aheads,
    ~ forecast_k_week_ahead(can_latest, "cr_7dav", "cr_7dav", .x, linear_reg())
  ),
  map(
    aheads,
    ~ forecast_k_week_ahead(
      can_latest, "cr_7dav", "cr_7dav", .x,
      boost_tree(mode = "regression", trees = 20)
    )
  )
)
```

The figures below shows the results for all of the provinces.

```{r plot-can-fc-lr, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 12}
ggplot(
  can_fc %>% filter(engine_type == "lm"),
  aes(x = target_date, group = forecast_date)
) +
  coord_cartesian(xlim = lubridate::ymd(c("2020-12-01", NA))) +
  geom_line(
    data = can_latest$DT %>% tibble(), aes(x = time_value, y = cr_7dav),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = geo_value),
    alpha = 0.4
  ) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 3) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(
    title = "Using simple linear regression", x = "Date",
    y = "Reported COVID-19 case rates"
  ) +
  theme(legend.position = "none")
```

```{r plot-can-fc-boost, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 12}
ggplot(
  can_fc %>% filter(engine_type == "xgboost"),
  aes(x = target_date, group = forecast_date)
) +
  coord_cartesian(xlim = lubridate::ymd(c("2020-12-01", NA))) +
  geom_line(
    data = can_latest$DT %>% tibble(), aes(x = time_value, y = cr_7dav),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = geo_value),
    alpha = 0.4
  ) +
  geom_line(aes(y = .pred)) +
  geom_point(aes(y = .pred), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 3) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(
    title = "Using boosted regression trees", x = "Date",
    y = "Reported COVID-19 case rates"
  ) +
  theme(legend.position = "none")
```

Both approaches tend to produce quite volatile forecasts (point predictions)
and/or are overly confident (very narrow bands), particularly when boosted
regression trees are used. But as this is meant to be a simple demonstration of
sliding with different engines in `arx_forecaster`, we may devote another
vignette to work on improving the predictive modelling using the suite of tools
available in `{epipredict}`.

</details>

## Version-aware forecasting

### Example using case data from US states

We will now employ a forecaster that uses properly-versioned data (that would
have been available in real-time) to forecast the 7 day average of future
COVID-19 case rates from current and past COVID-19 case rates and death rates
for all states. That is, we can make forecasts on the archive, `x`, and compare
those to forecasts on the latest data, `x_latest` using the same general set-up
as above. Note that in this example, we use a geo-pooled approach (using
combined data from all US states and territories) to train our model.

<details>

<summary>Download data using `{epidatr}`</summary>
```{r load-data, eval=FALSE}
# loading in the data
states <- "*"

confirmed_incidence_prop <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_incidence_prop",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200301, 20211231),
  geo_values = states,
  issues = epirange(20000101, 20221231)
) %>%
  select(geo_value, time_value, version = issue, case_rate = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_archive(compactify = FALSE)

deaths_incidence_prop <- pub_covidcast(
  source = "jhu-csse",
  signals = "deaths_incidence_prop",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200301, 20211231),
  geo_values = states,
  issues = epirange(20000101, 20221231)
) %>%
  select(geo_value, time_value, version = issue, death_rate = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_archive(compactify = FALSE)


x <- epix_merge(confirmed_incidence_prop, deaths_incidence_prop, sync = "locf")

x <- x %>%
  epix_slide(
    .versions = fc_time_values,
    function(x, gk, rtv) {
      x %>%
        group_by(geo_value) %>%
        epi_slide_mean(case_rate, .window_size = 7L) %>%
        rename(case_rate_7d_av = slide_value_case_rate) %>%
        epi_slide_mean(death_rate, ..window_size = 7L) %>%
        rename(death_rate_7d_av = slide_value_death_rate) %>%
        ungroup()
    }
  ) %>%
  rename(version = time_value) %>%
  rename(
    time_value = slide_value_time_value,
    geo_value = slide_value_geo_value,
    case_rate = slide_value_case_rate,
    death_rate = slide_value_death_rate,
    case_rate_7d_av = slide_value_case_rate_7d_av,
    death_rate_7d_av = slide_value_death_rate_7d_av
  ) %>%
  as_epi_archive(compactify = TRUE)

saveRDS(x$DT, file = "case_death_rate_archive.rds")
```

```{r load-stored-data}
x <- readRDS("case_death_rate_archive.rds")
x <- as_epi_archive(x)
```
</details>

Here we specify the ARX model.

```{r make-arx-model}
aheads <- c(7, 14, 21)
fc_time_values <- seq(
  from = as.Date("2020-09-01"),
  to = as.Date("2021-12-31"),
  by = "1 month"
)
forecaster <- function(x) {
  map(aheads, function(ahead) {
    arx_forecaster(
      epi_data = x,
      outcome = "death_rate_7d_av",
      predictors = c("death_rate_7d_av", "case_rate_7d_av"),
      trainer = quantile_reg(),
      args_list = arx_args_list(lags = c(0, 7, 14, 21), ahead = ahead)
    )$predictions
  }) %>%
    bind_rows()
}
```

We can now use our forecaster function that we've created and use it in the
pipeline for forecasting the predictions. We store the predictions into the
`arx_preds` variable and calculate the most up to date version of the data in the
epi archive and store it as `x_latest`.

```{r running-arx-forecaster}
arx_preds <- x %>%
  epix_slide(
    ~ forecaster(.x),
    .before = 120, .versions = fc_time_values
  ) %>%
  mutate(engine_type = quantile_reg()$engine) %>%
  mutate(ahead_val = target_date - forecast_date)

x_latest <- epix_as_of(x, version = max(x$versions_end))
```

Now we plot both the actual and predicted 7 day average of the death rate for
the chosen states

<details>

<summary>Code for the plot</summary>
```{r plot-arx-asof, message = FALSE, warning = FALSE}
states_to_show <- c("ca", "ny", "mi", "az")
fc_states <- arx_preds %>%
  filter(geo_value %in% states_to_show) %>%
  pivot_quantiles_wider(.pred_distn)

x_latest_states <- x_latest %>% filter(geo_value %in% states_to_show)

p2 <- ggplot(fc_states, aes(target_date, group = forecast_date)) +
  geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = geo_value), alpha = 0.4) +
  geom_line(
    data = x_latest_states, aes(x = time_value, y = death_rate_7d_av),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_line(aes(y = .pred, color = geo_value)) +
  geom_point(aes(y = .pred, color = geo_value), size = 0.5) +
  geom_vline(aes(xintercept = forecast_date), linetype = 2, alpha = 0.5) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 1L) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Date", y = "7 day average COVID-19 death rates") +
  theme(legend.position = "none")
```
</details>

```{r show-plot2, fig.width = 9, fig.height = 6, echo = FALSE}
p2
```
